---
title: "{{< var course.short >}} - Course Prediction Competition"
---

```{r}
#| echo: false
mtime_ods <- fs::file_info("key_dates.ods")$modification_time
mtime_csv <- fs::file_info("key_dates.csv")$modification_time
if(mtime_ods > mtime_csv){
  system("soffice --convert-to csv key_dates.ods")
}
```

```{r}
#| message: false
#| echo: false
library(tidyverse)
COMPETITION_START <- read_csv("key_dates.csv") |>
    filter(Item == "Prediction Competition") |>
    filter(str_detect(Details, "Opened")) |> 
    pull(Date) |> 
    strftime("%A %B %d, %Y")

COMPETITION_ENDS <- read_csv("key_dates.csv") |>
    filter(Item == "Prediction Competition") |>
    filter(str_detect(Details, "Closed")) |> 
    pull(Date) |> 
    strftime("%A %B %d, %Y")


REPORT_DUE <- read_csv("key_dates.csv") |>
    filter(Item == "Prediction Competition") |>
    filter(str_detect(Details, "Report")) |> 
    mutate(dt = as.POSIXct(paste(Date, "23:59pm"))) |> 
    pull(dt) |> 
    strftime("%A %B %d, %Y at %H:%M %p")
```

In lieu of a final exam, {{< var course.short >}} has a prediction competition
worth 40% of your final grade. The 400 total points associated with this 
competition will be awarded as:

- 300 points: Prediction Accuracy
- 100 points: Final report 

::: {.callout-important title="Use only Provided Information"}

You **may not** use any external data beyond what the instructor provides.
You will be required to attest to this in your final submission. 
Students found to use data outside instructor-provided files will receive
a zero on _all_ grades related to the course competition and will likely fail
the course as a result.

You may read published research about predictors of student success to gain
*qualitative* background information, but you may not access their data. 
You will be required to cite all background reading in your final report.

:::

## Background

Relevant background for the competition will be posted by `r COMPETITION_START`.

## Data

Data files for the competition will be posted by `r COMPETITION_START`.

## Example Predictions {#example_submission}

Example scripts illustrating the use of the training data to create predictions
and suitable formatting for Kaggle submission will be posted 
by `r COMPETITION_START`. 

## Grading

### Prediction Accuracy (200 Points)

Prediction accuracy will be assessed using *Kaggle Competitions*. I will
provided Kaggle with a private test set on which your predictions will be
evaluated. Kaggle will further divide the test set into a *public leaderboard*
and a *private leaderboard*. You will be able to see your accuracy on the
public leaderboard but final scores will be based on *private leaderboard*
accuracy so *don't overfit*. 

You will be allowed to submit only *two* sets of predictions to Kaggle per day
so use them wisely. At the end of the competition, you will be able to select
**two** sets of predictions to be used for your final evaluation (taking the
better of the two) so track your submissions carefully. 

**Kaggle will close for submissions at 11:55pm on `r COMPETITION_ENDS` 
for this class. Make sure your final submissions have been made in time.**

#### Grading

Grades for this portion will be assigned based on your performance relative
to both the instructor provided baselines and the best score obtained by
your fellow students.[^curve]

In particular, the top score in the class will get a score of 200 automatically.
More details about the scoring formula will be provided when the competition
opens on `r COMPETITION_START`.

[^curve]: Note that this is essentially the same as 'curving' the scores so that
the top-performing student automatically gets a perfect score and other grades
are raised accordingly. By judging against the best student score, rather than
an absolute benchmark like '0 error', all students' grades are implicitly 
raised.

### Final Report (100 Points)

By `r REPORT_DUE`, each competition unit (either an individual or a team) will
submit a final report of no more than 10 pages (10-12 point, one-and-a-half or
double-spaced) providing an *After-Action Report* of their competition 
performance. This report should focus on three topics: 

- 1) Are there any systematic errors in model predictions that need to be 
  addressed before this model could be applied broadly. (*E.g.*, is it 
  systmatically low or high in a particular subgroup; does it under-predict for 
  especially high outcome samples; etc?)
- 2) What insights into the underlying data can be gleaned from the model? 
     *E.g.*, are certain features especially important for making predictions? 
     Or are certain features which you would expect to be important not actually
     particularly important? 
- 3) What steps did your team take that were particularly helpful to maximizing
     predictive performance? Or, what parts of your model development cycle were
     weak and could be the most improved?[^selfeval] 

[^selfeval]: In this 'self-evaluation' section, you are encouraged to be 
truthful and honest in your reflections. I already know how well you did and 
you won't be able to convince me otherwise, so if you made fundamental errors 
that hindered your performance, I would rather see them discussed honestly 
(indicating understanding of how you could improve) rather than minimized. 

Note that this report is not solely focused on predictive performance. Analysis
that dives deep into the underlying structure of the data and generates
novel insights will score as well (or perhaps even better) than a highly 
performant but noninterpretable model. 

To assist in developing this After-Action Report, the instructor will provide
non-anonymized versions of the data (as well as a mapping to the anonymized data)
after the Kaggle competition ends. 

The report should include all code used to prepare the data and train and predict
from the best performing models in an Appendix. Significant penalties may be 
applied if the course staff is unable to reliably reproduce your predictions. 
(You may choose to submit this Appendix in the form of an iPython Notebook, 
Quarto document, Docker container, *etc* to maximize reproducibility.) This
appendix does not count against your 10 page limit.

You may, but are not required to, share your code with the instructor via
an emailed Zip file or link to a public code hosting platform such as GitHub.

::: {.callout-tip title="Reproducibility Info"}

To maximize the reproducibility of your code, make sure to: 

1) Avoid hard-coding any file paths. It is better to download and/or read
   directly from hosted copies whenever possible. 
2) Save random seeds used to create data splits, initialize training, *etc.*
3) List all software and packages used, *including version information.*
4) Have a clear set of 'reproduction steps' and accompanying documentation.

:::

Teams of two should both submit a copy of their final report. The instructor
will de-duplicate submissions.

The report will *roughly* be assessed acccording to the following rubric though
the instructor may deviate as necessary. 

::: {.list-table width="1,1,1,1,1,1" aligns="c,c,c,c,c,c" header-rows=1 .hover}

 * - Report Element
   - Excellent. <br> "A-" to "A" (`90% to 100%`)
   - Great. <br> "B-" to "B+" (`80% to 89%`)
   - Average. <br> "C" to "C+" (`73% to 79%`)
   - Poor. <br> "D" to "C-" (`60% to 72%`)
   - Failure. <br> "F" (`0% to 59%`)
 * - Quality of Report (15 points)
   - Excellent Report. Report has excellent writing and formatting, with 
     particularly effective tables and figures. Tables and Figures are
     "publication-quality" and clearly and succinctly support claims made in 
     the body text. Text is clear and compelling throughout. 
   - Great Report. Report has strong writing formatting. Text is generally clear,
     but has multiple minor weaknesses or one major weakness; tables and figures
     make their intended points, but do not do so optimally. 
   - Solid Report. Report exhibits solid written communication, key points are
     made understandably and any grammatical errors do not impair
     understanding. Code, results, and text could be better integrated, 
     but it is clear which elements relate. Formatting is average; tables and 
     figures do not clearly support arguments made in the text and/or are not
     "publication quality".
   - Poor Report. Written communication is below standard: points are not always
     understandable and/or grammatical errors actively distract from
     content. Code, results, and text are not actively integrated, but
     are generally located 'near' each other in a semi-systematic fashion.
     Poor formatting distracts from substance of report. Tables and
     Figures exhibit significant deficiencies in formatting.
   - Unacceptable Report. Written communication is far below standard, possibly
     bordering on unintelligible. Formatting prohibits or significantly impairs
     reader understanding.
 * - Analysis of Predictive Accuracy (10 points)
   - Excellent Analysis. Team is able to clearly identify strengths and 
     weaknesses of their model and to propose extensions / next steps that
     could use de-anonymized structure to further improve model performance. 
   - Great Analysis. Team identifies strengths and weaknesses of their model, 
     but without clear 'next steps' for model improvement.
   - Solid Analysis. Accuracy analysis successfully identifies patterns of 
     error, but does not connect these to modeling. 
   - Poor Analysis. Accuracy analysis attemps to identify patterns of error, 
     but fails to distinguish _systematic_ error from randomness. 
   - Unacceptable Analysis. Accuracy analysis is superficial and does not
     take advantage of data structure in a meaningful way. 
 * - Model-Driven Insights into Data Generating Process (10 points)
   - Excellent Insights. Modeling process creates significant new insights into
     the economics of real property assessment. Insights are then used to 
     further improve predictive modelling in a virtuous cycle. 
   - Great Insights. Modeling process creates significant new insights into
     the economics of real property assessment, but insights do not improve
     predictive modeling. 
   - Solid Insights. Modeling process creates new non-trivial insights, but 
     not ones that have major impact on predictive performance. (*E.g.*, 
     grey houses have a much higher chance of having rooftop solar than other
     house colors because grey was the most popular 'builder spec color' by
     the time that residential solar became commonplace. Interesting, but not
     especially helpful.)
   - Poor Insights. Modeling process only reproduces known / trivial insights
     about data generating process (*e.g.*, bigger houses are worth more than
     smaller houses *ceteris paribus*).
   - Unacceptable Insights. No attempt is made at generating meaningful insights
     from models. 
 * - Reflection on Competition Workflow (10 points)
   - Excellent Reflection. Clear identification of all important good and bad
     decisions made over the course of the competition, with insightful 'take 
     aways' that can be used by self and other teams to significantly improve
     performance on future prediction tasks. Importance of key decisions is
     clearly demonstrated. 
   - Great Reflection. Impressive reflection on key decisions (good and bad)
     made over the course of the competition. 'Take Away' messages would be 
     useful if this competition were re-run as is (or with minor changes) but
     do not necessarily generalize to other similar tasks. Importance of key
     decisions is partially demonstrated. 
   - Solid Reflection. Reflection on key decisions identifies major 
     choices made throughout competition, but fails to fully analyze
     their impact. 'Take away' messages are useful, but generic and not 
     particularly relevant to this course or this competition. (*E.g.*, advice
     on the best way to tune the lasso) Minimal effort to demonstrate importance
     of key decisions. 
   - Poor Reflection. Reflection seems to miss one or more major choices made
     over the course of the semester OR attributes too much importance to an 
     unimportant decision. Fails to demonstrate importance of key decisions.
     'Take away' messages are of limited general applicability.
   - Unacceptable Reflection. Minimal or shallow reflection. 'Take away' 
     messages are trivial or misleading. 
 * - Reproduction Code (5 points)
   - Excellent Reproduction Code. Code is easy to read and execute, with excellent
     commenting, formatting, *etc.* and clearly reproduces submitted predictions.
   - Great Reproduction Code. Code is easy to read, but requires some effort
     to execute and reproduce submitted predictions. 
   - Solid Reproduction Code. Code lacks clarity, but still appears to reproduce
     submitted predictions with reasonable effort. 
   - Poor Reproduction Code. Instructor cannot reproduce submitted predictions
     without significant effort. 
   - Unacceptable Reproduction Code. Code cannot reproduce submitted predictions.
   
:::

At the end of your report, you **must** include a detailed description of the 
extent to  which you used Generative AI tools in this course competition or in
preparation of your final report. Failure to include an AI disclosure will 
result in an **automatic 25% penalty.**

Recall that Generative AI **may not** be used for writing your final report;
it may only be used for coding assistance.

## Teams

Students *may*, but are not required to, work in teams of 2 for this competition.
Teams will submit one (shared) final presentation and final report and will 
receive the same grade for **all course elements**. Students will create teams
*via* Kaggle's team functionality. 

Teams may be formed at any point in the competition. Teams will not be allowed
to dissolve except under truly exceptional circumstances. (Unequal participation
is sadly not an exceptional circumstance.)
