---
title: "{{< var course.short >}} - {{< var course.title >}}"
format: "html"
---

```{r}
#| echo: false
mtime_ods <- fs::file_info("key_dates.ods")$modification_time
mtime_csv <- fs::file_info("key_dates.csv")$modification_time
if(mtime_ods > mtime_csv){
  system("soffice --convert-to csv key_dates.ods")
}
```

Welcome to the course website for {{< var course.short >}}  - 
{{< var course.title >}} ({{< var course.semester >}})!

{{< var course.short >}} is a rigorous, but non-mathematical introduction to
classical and modern machine learning techniques, with a focus on their 
statistical foundations. Building on a foundation of classical linear and 
logistic regression models, core principles of modern statistical learning will 
be covered: cross-validation, resampling methods, regularization, model 
selection, ridge and lasso regression, margin-based classifiers, semi-parametric
and kernel methods, tree-based methods, ensemble methods and neural networks. 
The primary focus of this course will be on the implementation and application 
of these methods, not formal theory. 

Students will complete the course with a strong conceptual understanding the 
fundamentals of statistical machine learning and a practical understanding of 
many of the most useful methods of modern machine learning. 

On this site, you will find useful material including: the 
[Course Syllabus](./syllabus.qmd), [Problem Sets](./problem_sets.qmd), 
[Instructor Notes and Handouts](./notes.qmd), and details of the 
[Applied Final Exam (Course Competition)](./competition.qmd). Copies of each
of these documents can be found on our Courseworks page. 

There are quite a few moving parts to this course, so this [key dates file](./key_dates.csv)
or the list of upcoming course activities below may be useful:

```{r}
#| echo: false
#| message: false
#| warning: false
library(yaml)
library(glue)
library(tidyverse)
if(file.exists("_variables.yml")){
    variables <- read_yaml("_variables.yml")
} else {
    variables <- read_yaml("https://raw.githubusercontent.com/michaelweylandt/STAT3106/refs/heads/main/_variables.yml")
}
course_short <- variables$course$short

read_csv("key_dates.csv") |> 
    filter(Item %in% c("Class Session", 
                       "Homework", 
                       "No Class", 
                       "Prediction Competition")) |>
    group_by(Item) |> 
    mutate(Details = case_when(
      Item == "Class Session" ~ paste0("Lecture #", row_number(), ": ", Details), 
      TRUE ~ Details)) |>
    ungroup() |> 
    arrange(Date) |> 
    mutate(Date = strftime(Date, "%a %b %d")) |>
    select(-Item, -Number, -`Day of Week`) |>
    DT::datatable(options=list(searching=FALSE,
                               paging=TRUE,
                               lengthChange=FALSE,
                               ordering=FALSE),
                  style='bootstrap',
                  caption=glue("Upcoming Important Dates for {course_short}"))
```

<script defer>
window.addEventListener("load", function(){
  var filter_date = new Date();
  filter_date.setDate(filter_date.getDate() - 1);
  var filter_string = filter_date.toISOString().split('T')[0]
  $("table").DataTable().rows((idx, data) => data[4] < filter_string).remove().draw();
});
</script>

```{r}
#| message: false
#| echo: false
#| warning: false
library(tidyverse)
## Generate Google Calendar Formatted CSV for Student Events
DATES <- read_csv("key_dates.csv")

ROOM <- "Hamilton Hall 602"
START <- "10:10 AM"
END <- "11:25 AM"

library(yaml)
location <- read_yaml("_variables.yml")$course$location
course   <- read_yaml("_variables.yml")$course$short

`%rep_na%` <- function(x, y) ifelse(is.na(x), y, x)

### i) Lectures
LECTURES <- DATES |>
    filter(Item == "Class Session") |>
    mutate(Subject=paste0(course, "Lecture", Number, ": ", Details),
           `Start Date` = Date,
           `Start Time` = START,
           `End Date` = Date,
           `End Time` = END,
           `All Day Event` = "False",
           `Location` = ROOM) |>
    select(Subject, starts_with("Start"), starts_with("End"), `All Day Event`, Location)

### ii) Assignments
ASSIGNMENTS <- DATES |>
    filter(Item == "Homework") |>
    mutate(Subject = paste0(course, ": ", Details),
           `Start Date` = Date,
           `Start Time` = if_else(str_detect(Details, "Due"), "4:45 PM", NA),
           `End Date` = Date,
           `End Time` = if_else(str_detect(Details, "Due"), "5:00 PM", NA),
           `All Day Event` = if_else(str_detect(Details, "Released"), "True", "False"),
           `Location` = if_else(str_detect(Details, "Due"), "Gradescope", NA)) |>
    select(Subject, starts_with("Start"), starts_with("End"), `All Day Event`, Location)


### iii) Course Competition Elements
COMPETITION <- DATES |>
    filter(Item == "Prediction Competition") |>
    mutate(Subject=paste0(course, ": ", Details),
           `Start Date`= Date,
           `Start Time`= if_else(str_detect(Details, "Due"), "11:30 PM", NA),
           `End Date`= Date,
           `End Time`= if_else(str_detect(Details, "Due"), "11:55 PM", NA),
           `All Day Event` = if_else(str_detect(Details, "Due"), "False", "True"),
           Location = if_else(str_detect(Details, "Due"), "Gradescope", NA)) |>
    select(Subject, starts_with("Start"), starts_with("End"), `All Day Event`, Location)

### iv) Midterm
MIDTERM <- DATES |>
    filter(Item == "Midterm Exam") |>
    mutate(Subject=paste(course, "Midterm Exam: ", Details),
           `Start Date` = Date,
           `Start Time` = START,
           `End Date` = Date,
           `End Time` = END,
           `All Day Event` = "False",
           `Location` = ROOM) |>
    select(Subject, starts_with("Start"), starts_with("End"), `All Day Event`, Location)

bind_rows(LECTURES, ASSIGNMENTS, COMPETITION, MIDTERM) |>
    arrange(`Start Date`) |>
    write_csv("key_dates_students_gcal.csv")
```

A CSV file suitable for import into Google Calendar with all assignment deadlines
can be found [here](./key_dates_students_gcal.csv).

Instructor: [Michael Weylandt](https://michael-weylandt.com)
