---
title: "STAT UN3106\nApplied Machine Learning"
date: "`r Sys.Date()`"
language: 
  title-block-published: "Last Updated"
format:
  html:
    email-obfuscation: javascript
---

```{r}
#| echo: false
mtime_ods <- fs::file_info("key_dates.ods")$modification_time
mtime_csv <- fs::file_info("key_dates.csv")$modification_time
if(mtime_ods > mtime_csv){
  system("soffice --convert-to csv key_dates.ods")
}
```

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(glue)

monday_date <- function(week=WEEK){
  starting_monday <- read_csv("key_dates.csv") |> 
    pull(Date) |> 
    min() |>
    floor_date("week", week_start=1)
  
  starting_monday + 7 * (week - 1)
}

wednesday_date <- function(week=WEEK){
  monday_date(week) + 2 
}

friday_date <- function(week=WEEK){
  monday_date(week) + 4
}

wednesday_topic <- function(week=WEEK){
  read_csv("key_dates.csv", show_col_types=FALSE) |>
    filter(Date == wednesday_date(week)) |>
    pull(Details) |>
    coalesce("")
}

wednesday_reading <- function(week=WEEK){
  read_csv("key_dates.csv", show_col_types=FALSE) |>
    filter(Date == wednesday_date(week)) |>
    pull(`Suggested Reading`) |>
    coalesce("")
}


monday_topic <- function(week=WEEK){
  read_csv("key_dates.csv", show_col_types=FALSE) |>
    filter(Date == monday_date(week)) |>
    pull(Details) |>
    coalesce("")
}

monday_reading <- function(week=WEEK){
  read_csv("key_dates.csv", show_col_types=FALSE) |>
    filter(Date == monday_date(week)) |>
    pull(`Suggested Reading`) |>
    coalesce("")
}

hw <- function(week=WEEK){
  hw_text <- read_csv("key_dates.csv", show_col_types=FALSE) |>
    filter(Date == friday_date(week)) |> 
    pull(Details) 
  
  if(NROW(hw_text) == 0) return("")
  
  hw_number <- str_extract(hw_text, "[:digit:]+")
  
  glue("[{hw_text}](problem_sets/ps0{hw_number}.qmd) due on {strftime(friday_date(week), '%Y-%m-%d')} at 5:00pm")
  
}

class_date <- function(n, details=FALSE){
    c <- read_csv("key_dates.csv") |>
        filter(`Course Element` == "Class Session") |>
        filter(`Item Number` == n)
    
    if(details){
        c |> pull(Details)
    } else {
        c |> pull(Date) |>
             as.Date() |>
             format("%Y-%m-%d")
    }
}

report_date <- function(){
  read_csv("key_dates.csv") |>
        filter(str_detect(Details, "Report Due")) |>
        pull(Date)
}


grade_submission_date <- function(){
  read_csv("key_dates.csv") |>
        filter(str_detect(Details, "Grade Submission")) |>
        pull(Date)
}
```

**All syllabus and schedule provisions subject to change with 
suitable advance notice. Changes will be announced in class and posted 
on Canvas.**

## Course Description

*Applied Machine Learning* (`STAT UN3106`) is a lecture course meant to follow
*Introduction to Applied Data Science* (`STAT UN2108`).  Building on a foundation
of classical linear and logistic regression models, core principles of modern
statistical learning will be covered: cross-validation, resampling methods, 
regularization, model selection, ridge and lasso regression, margin-based 
classifiers, semi-parametric and kernel methods, tree-based methods, ensemble 
methods and neural networks. The primary focus of this course will be on the
implementation and application of these methods, not formal theory. 

Students will complete the course with a strong conceptual understanding the 
fundamentals of statistical machine learning and a practical understanding of 
many of the most useful methods of modern machine learning. 

## Course Staff

### Instructor

[Michael Weylandt](https://michael-weylandt.com)

- Office Hours: TBD
- Location: TBD
- Email: <rmw2185@columbia.edu>

### Teaching Assistant(s)

[Lenny Aharon](https://lennyaharon.github.io/index.html)

- Office Hours: Thursdays 8:30-9:30am 
- Location: Virtual (Zoom - Link Distributed via Courseworks)
- Email: <lenny.aharon@columbia.edu>

### Asynchronous Support

In addition to office hours, asynchronous support will be available via
the course Ed Discussion board. The Course Staff will endeavor to reply to
student posts within **two business days**, so plan to ask questions well in 
advance of any tests or homework due dates. 

## Textbooks & Course Materials

The primary textbook for this course will be *An Introduction to Statistical 
Learning with Applications* (ISL) by Gareth James, Daniela Witten, 
Trevor Hastie, Robert Tibshirani, and Jonathan Taylor. There are two versions of
this textbook freely 
available online at <https://www.statlearning.com>: 

- *An Introduction to Statistical Learning with Applications in R* (ISLR)
- *An Introduction to Statistical Learning with Applications in Python* (ISLP)

The two versions cover the same material, only using different languages for the
examples, so please feel free to use whichever version you find more comfortable.

Additional readings will be assigned by the instructor from the following
texts, all of which are freely available online: 

- [*Patterns, Predictions, and Actions: A story about machine
  learning*](https://mlstory.org/) by Moritz Hardt and Benjamin Recht (PPA)
- [*Fairness and Machine Learning: Limitations and
  Opportunities*](https://fairmlbook.org/) by 
  Solon Barocas, Moritz Hardt, and Arvind Narayanan (FML)
- [*The Emerging Science of Machine Learning 
  Benchmarks*](https://www.mlbenchmarks.org/) by Moritz Hardt (MLB)
  
## Grading

Modern statistical learning requires mastery on multiple levels: theory, 
methodology, and applications. While the main focus of this course lies in
applications, the grading in this course is designed to evaluate students on
all three elements. Specifically, your grade depends on three major components

- 30% In-Class Midterm Exam
- 30% Homework (4, Equally Weighted)
- 40% Applied Final Exam (Prediction Contest)

Points will be aggregated for a maximum class score of 1000. Final grades
will be curved upwards in accordance with relevant program, departmental,
and college policies. (Final grades will not be curved downwards.) Roughly,
the following portion of final grades are expected: 

- A Range (A to A-): <= 40%
- B Range (B+ to B-): <= 40%
- C- and Lower: <20% 

### In-Class Midterm (Theory and Concepts)

One in-class test, worth 300 points, will be given approximately two-thirds of
the way through the semester. Each tests will have a mixture of multiple choice, 
short answer, and long-answer questions covering material covered in lecture, 
instructor notes, and assigned readings.

Rules to avoid opportunities for cheating will be distributed in class before 
the midterm.  Please review the rules and let your instructor know before the 
exam if you need an accommodation regarding the rules.  There will be zero 
tolerance for failure to abide by those rules.

No outside materials will be allowed during the tests. If necessary, the 
instructor will provide a sheet of key definitions and formulas. 

### Homework (Methodology)

{{< include _homework_syllabus.qmd >}}

### Applied Final Exam (Applications)

In lieu of a traditional final exam, {{< var course.short >}} has a 
semester-long "prediction competition" in which students will develop, tune, 
and deploy machine learning algorithms on a difficult instructor-provided 
data set. Your grade in this competition will be based on three components: 

- 200 Points: Prediction Accuracy
- 100 Points: In-Class Presentation
- 100 Points: Final Report

#### Prediction Accuracy

Prediction accuracy will be assessed using a private *Kaggle Competition*. 
The instructor will provide Kaggle with a private test set on which your
predictions will be evaluated. Kaggle will further divide the test set into a
*public leaderboard* and a *private leaderboard*. You will be able to see your 
accuracy on the public leaderboard but final scores will be based on private 
leaderboard accuracy *so don't overfit*. 

You will be allowed to submit only two sets of predictions to Kaggle per day so
use them wisely. At the end of the competition, you will be able to select two
sets of predictions to be used for your final evaluation (taking the better of 
the two) so track your submissions carefully.

Further details about the data set, prediction metrics and grading scheme will 
be provided later in the course. 

#### In-Class Presentation

During the final full week of class[^week], students will give a **6** minute
presentation, primarily focused on the *models and methods* they used to build
their predictive model. Details of presentations, including the presentation
schedule, will be made available later in the course. 

**Students will present in an order established by the instructor and should 
expect to attend course on both presentation days.**

If you have a foreseeable circumstance that requires you to present on a 
particular date, please contact the instructor or the TA well in advance.
Failure to present on the assigned day will incur an automatic 0 on this portion
of your grade. 

[^week]: If it turns out to be possible to have all students present in a single
class period, presentations will be on the *Wednesday* and Monday will be used
for a special topics lecture. This will be announced as we get closer to the
end of the semester.

#### Final Report

After the competition ends, you will submit a final report of no more than 8
pages[^limit] (10-12 point, one-and-one-half- or double-spaced) providing an 
*After-Action Report* of your competition performance. This report should 
focus on three topics:

[^limit]: Any pages after the first 8 will be discarded and not graded.

- Are there any systematic errors in model predictions that need to be addressed
  before this model could be applied broadly. (*E.g.*, is it systematically low 
  or high in a particular subgroup; does it under-predict for especially high
  outcome samples; *etc*?)
- What insights into the underlying data can be gleaned from the model? *E.g.*,
  are certain features especially important for making predictions? Are certain
  features which you would expect to be important not actually particularly
  important?
- What steps did you take that were particularly helpful to maximizing 
  predictive performance? Or, what parts of your model development cycle were 
  weak and could be the most improved?

Note that this report is not solely focused on predictive performance. Analysis
that dives deep into the underlying structure of the data and generates novel
insights will score as well (or perhaps even better) than a highly performant
but noninterpretable model.

To assist in developing this After-Action Report, the instructor will provide
non-anonymized versions of the data (as well as a mapping to the anonymized 
data) after the Kaggle competition ends.

The report should include all code used to prepare the data and train and 
predict from the best performing models in an Appendix. Significant penalties
will be applied if the instructor is unable to reliably reproduce your
predictions. (You may choose to submit this Appendix in the form of an iPython
Notebook, Quarto document, Docker container, *etc.* to maximize reproducibility.)
Note that this appendix does not count against your 8 page limit. You will 
submit your appendix either via an emailed Zip archive or a link to a public
code hosting platform such as GitHub. 

## Course Policies

### Disability Accommodations

Students should review the material at 
<https://www.college.columbia.edu/rightsandresponsibilities>. If you believe 
that you might have a disability that requires accommodation, you should contact
Disability Services (Columbia) or CARDS (Barnard).

### Academic Integrity

Students should review Columbia College's Academic Integrity standards at
<https://www.cc-seas.columbia.edu/academic-integrity> and consult with the
instructor if they have any questions. 

Problem Sets: Students are allowed (and encouraged!) to work with each other
on problem sets. Additionally, students may use *free* tiers of 
generative AI software or any other free web resources to complete any aspect 
of the problem sets. Students will be required to list all resources used 
(including generative AI tools or classmates) at the end of each problem set. 

Tests: No external resources may be used.

Applied Final Exam: Students are allowed to work with each other on the 
prediction competition (but keep in mind that this is a *competition* and your 
grade is based on your outcome relative to your classmates). Additionally, 
students may use *free* tiers of generative AI software or any other free web 
resources to develop and implement predictive models. Generative AI may not be
used to write the final report. Students will be required to list all resources 
used(including generative AI tools or classmates) as part of their final submission.

**Punishment for violations of relevant academic integrity standards range from
a 50% penalty on the assignment to an automatic failure of the course. 
In matters of academic integrity, it is *far* better to ask permission than
forgiveness.**

## Tentative Schedule

Class sessions will alternate between traditional "lecture" sessions, focused on
key concepts and methods in machine learning, and "lab" sessions, focused on
application and interpretation of methods to real data sets. This is not a hard
distinction: some applications may be demonstrated in "lecture" sessions and
ideas will be introduced and reviewd in "lab" sessions.

For each class session, suggested "pre-reading" will be assigned from one of
the course textbooks, typically ISLR/ISLP. This is optional, but strongly 
recommended, for students who want to have prior exposure to the fundamentals
of that day's topic before lecture.

::: {.list-table width="1,1,3,2,1, 3,2,1" aligns="c,l,l,l,l,l,l,l" header-rows=1 .hover}

  * - Week 
    - Monday Date
    - Monday Topic
    - Monday Pre-Reading
    - Wednesday Date
    - Wednesday Topic
    - Wednesday Pre-Reading
    - Homework

  * - `r (WEEK <- 1)`
    - 
    - 
    - 
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()` 

  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - `r wednesday_date()`
    - `r wednesday_topic()`
    - `r wednesday_reading()`
    - `r hw()`
    
  * - `r (WEEK <- WEEK + 1)`
    - `r monday_date()`
    - `r monday_topic()`
    - `r monday_reading()`
    - 
    - 
    - 
    - `r hw()`

  * - 
    - Finals Period
      <td colspan="6" style="text-align:center"> 
        <p>Final Report due `r report_date()` (Tentative)</p> 
        </br> 
        <p>Final Grades posted by `r grade_submission_date()`</p>
      </td>
:::


**All syllabus and schedule provisions subject to change with 
suitable advance notice. Changes will be announced in class and posted 
on Canvas.**
