
```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
get_ps_assigned <- function(N=ps){
    read_csv("key_dates.csv") |> 
        filter(Item == "Homework", 
               Number == N, 
               str_detect(Details, "Released")) |>
        pull(Date) |>
        format("%A %B %d")
}

get_ps_due <- function(N=ps){
    read_csv("key_dates.csv") |> 
        filter(Item == "Homework", 
               Number == N, 
               str_detect(Details, "Due")) |>
        pull(Date) |>
        format("%A %B %d at 5:00PM on Canvas/Courseworks")
}
```

- Problem Set Assigned: `r get_ps_assigned(ps)`
- **Due Date: `r get_ps_due(ps)`**

::: {.callout-important title="Submission Instructions"}

Submit your assignment as a PDF on Canvas/Courseworks. 

Please observe the following guidelines when submitting an assignment: 

1) Use 10-12 point font, 1.25 or 1.5-spaced. 
2) Each problem should start on a new page. It is allowed, but not required,
   for subparts to start a new page, *e.g.*, you do not need a page break before
   starting 2(b), but you do need a page break before starting 2(a).
3) When including code alongside a submission, run it through a code
   formatting tool and keep lines to no more than 80 characters so they don't 
   run off the end of the page.  Unintelligible code is presumtively incorrect.
4) Figures and tables should have proper titles, axis labels, units, *etc.*

You can use any formatting engine you prefer (`jupyter`, `R Markdown`, *etc*), 
but I recommend [`Quarto`](https://quarto.org/). 

Examples: [`R`-focused](./ps_example_r.qmd) and [`python`-focused](./ps_example_py.qmd)

:::

```{r}
#| include: false
fs::file_copy("problem_sets/ps_example_r.qmd", 
              "docs/problem_sets/ps_example_r.qmd")

fs::file_copy("problem_sets/ps_example_py.qmd", 
              "docs/problem_sets/ps_example_py.qmd")
```

$$\newcommand{\bx}{\mathbf{x}} \newcommand{\V}{\mathbb{V}} \newcommand{\E}{\mathbb{E}}$$
