---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}

**TODO: Add more examples here**

Outline: 

- Fitting OLS with Standard Software
- Computing Training and Test Error
- Beating The Best: Overfitting
- Using and Abusing Simulation Studies
- Real Data Example
- Other Loss Functions

--- 

## Review

Last time, we built up to the following 'meta-rule' for machine learning: 

$$\hat{f} = \argmin_{f \in \Fcal} \frac{1}{n}\sum_{i=1}^n L(f(\bX_i), Y_i)$$

Make sure you can identify what each of these terms mean! Specifically, 
you should be able to identify the _loss function_, the _predictor_, 
the _empirical risk_, the _model class_, and the _minimizer_. 

Recall that our overall strategy for machine learning is this: 

- Consider a class / family of models
- Try to have enough data, and sufficiently representative data, that
  past performance (training error) is a reasonable proxy for future
  performance (test error) within that model class
- In the absence of other information, pick the model that has the best
  past performance (empirical risk minimization) and hope that it will 
  do well going forward
- _Ceteris paribus_, a larger model class has more upside (best case better)
  and more downside (worse case worse) than a simpler model class. 
  With enough data, we hope to consistently find the 'good' parts of the
  bigger model family and to avoid the 'bad' parts. 
  

## Fitting OLS

Recall that Ordinary Least Squares (OLS) is the intersection of
empirical risk minimization (ERM), the class of linear models,
and the squared error loss. Formally: 

$$\hat{f}_{\text{OLS}} = \argmin_{f \in \Fcal_{\text{Linear Regression Models}}} \frac{1}{n}\sum_{i=1}^N (y_i - f(\bx_i))^2$$

where $\{(\bx_i, y_i)\}_{i=1}^n$ are the training data. If we
recall that all linear regression models ($\Fcal_{\text{Linear Regression Models}}$)
can be identified with the set of feature weights (coefficients), this
is equivalent to solving the problem: 

$$\hat{\bbeta} = \argmin_{\bbeta \in \R^p} \frac{1}{n}\sum_{i=1}^N (y_i - \bx_i^{\top}\bbeta)^2$$

where $\bbeta = (\beta_1, \beta_2, \dots, \beta_p)$ is a *vector* of regression
weights and $\bbeta^{\top}\bx = \beta_1x_1 + \beta_2x_2 + \dots + \beta_px_p$.

Computers don't work so well with the first format - a function is a bit of a
mathematical abstraction - but the second format (solving for a set of specific
numbers) is very easy for computers, so this is what gets used in practice, but
it is functionally just a different way of expressing the same problem. 

In fact, solving this problem - the _least squares_ problem - is ubiquitous in
mathematics, so _lots_ of work has been done on the best way to solve this and
we can benefit from this. In `R`, we use the `lm()` built-in, while `Python` 
gives us `np.linalg.solve()` among others. `R`'s `lm()` provides several
'quality-of-life' improvements that `np.linalg.solve()` lacks, so we'll start
with some `R` examples before coming back to `Python`. 

To demonstrate this, I've uploaded a 
[small `csv` file](./data/tx_power_bills.csv) with two columns: 

- Usage, the amount of electricity (measured in kWh) used by a single family 
  home over time periods lasting (roughly) one month each
- Bill, the bill (in US dollars) for that same time period

```{r}
library(tidyverse)
TX_POWER <- read_csv("notes/data/tx_power_bills.csv")

ggplot(TX_POWER, aes(x=Usage, y=Bill)) + geom_point() + 
    xlab("Power Usage (kWh)") + 
    ylab("Monthly Bill ($)") + 
    theme_bw() + 
    scale_y_continuous(labels=scales::dollar)
```

From background knowledge, you might expect this to be linear
(roughly $\text{Cost} = \text{Usage} * \text{Rate}) and indeed it
is _roughly_ linear, but not perfectly so.[^1] Despite this, we can
fit a line using the `lm()` function: 

```{r}
tx_power_model <- lm(Bill ~ Usage, data=TX_POWER)
tx_power_model
```
As an economic matter, this model is immediately a bit implausible: 
the negative intercept, if taken too literally, would imply that if
we used no power, the utility company would pay us \$3.90. Why is
this happening? 

Statisticians would usually say something like "This data is a bit 'noisy' / 
random" and that causes error in the model. It is true that there are many 
factors that impact billing beyond just plain usage. Here we have noise 
(or error) that arises not because the underlying process is inherently 
stochastic, but because there are aspects we don't include in our model. Whether
we should think about about this as pure 'noise' or as something a bit more 
systematic is a question we will return to later in the course. 

For now, let's start by making predictions from this model. Our model was
designed to have minimal training error, which we can compute directly: 

```{r}
MSE <- function(y_hat, y){
  mean((y_hat - y)^2)
}

y_hat <- predict(tx_power_model, data=TX_POWER)

MSE(y_hat, TX_POWER$Bill)
```

So our _average_ squared prediction error is 217.17! This is not great, but 
recall that the units here are "squared dollars", so it's a bit unnatural. 
We get a better intuition by computing the _root mean squared error_ instead: 

```{r}
RMSE <- function(y_hat, y){
  sqrt(MSE(y_hat, y))
}

RMSE(y_hat, TX_POWER$Bill)
```

Ok - so we're off by about \$14.74 each month. Certainly not perfect, but given
that the 'baseline' standard deviation of power bills is \$61.69, we've 
certainly achieved some predictive power. 

OLS is supposed to give us the predictions with the _best training error_, but
how can we verify this? If we don't trust the math, we can try out a wide range
of coefficients and see what actually gives the best performance. To do this, 
we need a bit of brute force calculations: 

```{r}
library(ggplot2)
compute_rmse <- function(intercept, slope){
  y_hat <- intercept + slope * TX_POWER$Usage
  RMSE(y_hat, TX_POWER$Bill)
}

# Create a 'grid' of many potential coefficients
potential_coefficients <- expand_grid(intercept=seq(-10, 10, length.out=401), 
                                      slope=seq(0, 3, length.out=401))

# Compute RMSEs for all coefficient combinations and plot results
potential_coefficients |> 
  mutate(training_rmse = map2_dbl(intercept, slope, compute_rmse)) |>
  ggplot(aes(x=intercept, y=slope, fill=training_rmse)) + geom_tile()
```
}

[^1]: Several reasons: inputs to the power grid change in price,
aggregate demand changes over time and prices balance out supply and
demand, 'base rate' charges increase over time to maintain the grid, 
*etc.*
