---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}

Outline: 

- Review of 'Statistical Learning' Definition
- Placing OLS within our ML Taxonomy
- Empirical Risk Minimization
- Review of OLS
- The Role of Model Classes

--- 

## A Formal Definition of Statistical Machine Learning

Recall that we posited this formal definition of _statistical learning_: 

> "The goal of machine learning is to find a function  $f: \Xcal \to \Ycal$
  that minimizes the risk 
  $R(f) = \E_{(\tilde{X}, \tilde{Y}) \sim \Dcal}[L(f(\tilde{X}), \tilde{Y})]$
  given only a sample $\{(X_i, Y_i)\}_{i=1}^n$ of observations from an
  unknown joint distribution $\Dcal$" 
  
That's a bit of a math-full mouthful, so let's start working through it. 
This definition has several interesting properties worth noting: 

- It seems (reasonably) actionable. While finding the best possible $f$ can be
  hard in practice, the idea of finding 'the best one' can be made workable
  in theory pretty easily, as we will see below. 
- It centers the role of _prediction_, possibly distinct from modeling. As far
  as the definition above is concerned, $f(\cdot)$ can be any function, whether
  it is physically plausible or not. Notably, this definition does not require
  $f(\cdot)$ to capture any sort of 'realistic' or 'causal' relationship. 
- It is essentially restricted to _supervised_ learning problems, in which our
  goal is to to predict a well-defined $Y$ from a given $X$. 
- We focus our predictions on _new_ data ('test data' $(\tilde{X}, \tilde{Y})$),
  not on just fitting old data that we have already seen 
  ('training data' $\{(X_i, Y_i)\}_{i=1}^n$) as well as possible. 
  
Our goal today is to think about how _Ordinary Least Squares_ (OLS) relates to
this definition. In doing so, we'll hopefully unpack the rather-mathy definition
above and make it a bit more easily understandable in a more familiar context.

## Loss Functions

Let's start by taking a closer look at the mathematical part of our goal: 

$$R(f) = \E_{(\tilde{X}, \tilde{Y}) \sim \Dcal}[L(f(\tilde{X}), \tilde{Y})]$$

Here, the loss function $L(\cdot, \cdot)$ is a measure of how accurate our 
predictions are. Note that $L$ takes two inputs - our prediction $\hat{Y}=f(X)$
and the right answer $Y$. Which _loss_ function should we choose? 

For sanity, we typically require that i) $L(y, y) = 0 \text{ for all } y$, *i.e.*,
the loss is 0 if our prediction is exactly correct; ii) $L$ generally
increases in $|y - \hat{y}|$, *i.e.*, our loss is larger when our guess is
'farther' from the truth. As a consequence of these two, $L$ is typically
non-negative ($L \geq 0$), so we never have 'negative loss'. Beyond these
basics, there are _many_ potential loss functions and they are often chosen for 
some combination of i) mathematical and computational ease of use; ii) inertia; 
and iii) relevance to the task at hand. 

We will discuss several of these loss functions over the semester, but the one
you are almost certainly most familiar with is _mean squared error_ (or its
cousin _root mean squared error_). Here, we take squared error for a single 
point: 

$$L(\hat{y}, y) = (\hat{y} - y)^2$$

and extend it 'pointwise' to larger data sets: 

$$\begin{align*} 
L &= \frac{1}{n}\sum_{i=1}^n (\hat{y}_i - y_i)^2 \tag{MSE} \\
L &= \sqrt{\frac{1}{n}\sum_{i=1}^n (\hat{y}_i - y_i)^2} \tag{RMSE}
\end{align*}$$

Here, we are applying $L$ to a series of predictions 
$\hat{y}_1, \hat{y}_2, \dots \hat{y}_n$ and comparing against (different) true 
values $y_1, y_2, \dots, y_n$. Note that the choice of MSE vs RMSE doesn't
really matter if we're just comparing two possible sets of predictions: if one 
set of predictions $\{\hat{y}_i\}_{i=1}^n$ has a better
MSE than another $\{\hat{y}'_i\}_{i=1}^n$, it will have a better RMSE as well
(and *vice versa*). The MSE/RMSE distinction is more one of interpretation and
scale, parallel to standard deviation/variance. 

### Means, Variance, MSE, and RMSE

You might have seen the _mean_ or _expectation_ of a random variable defined
using some sort of probability-weighted sum or integral. I'd like to offer
a different characterization that will be more helpful for building your ML
definition. 

Suppose you have a random variable $Z$ and you have to make the 'best' 
unconditonal prediction of $Z$. (By unconditional, I mean you have no 
ancillary information and you're just making a broad / uniform / constant
prediction.) If you are being evaluated under a squared error loss, 
the _mean_ of $Z$ is the single value that will give you the best score
on average. That is,[^p1]

$$\E[(Z - \mu_Z)^2] \leq \E[(Z - a)^2] \text{ for any } a \in \R$$

(Note that this same characterization applies to a data set, not just a full
random variable: if $x_1, \dots, x_n$ are known data points, the best prediction
against an opponent who will pick one of the $x_i$ uniformly at random is the
sample mean $\overline{x}$.[^emp])

So the _mean_ is the value that minimizes (expected) squared error. But how 
good will our predictions actually be if we use the mean as our predictor? That
is, what is $\E[(Z - \mu_Z)^2]$? Why, that's the variance of course!

This gives a nice little set of identities: if asked to predict a random
variable and evaluated under squared error loss: 

- The _mean_ is the loss-minimizer, *i.e.*, the best predictor
- The _variance_ is the minimum-MSE-loss, *i.e.*, the MSE that you'd get with 
  the best predictor
- The _standard deviation_ is the minimum-RMSE-loss, *i.e.*, the RMSE that you'd
  get with the best predictor

[^p1]: To see this, define $f(a) = \E[(Z - a)^2]$ and take a derivative to
get $f'(a) = \E[2(Z-a)] = 2\E[Z] - 2a$. If we set this to zero, we clearly get
$a = \E[Z]$ as the critical point for our predictions. 

[^emp]: This is an instance of a more general point of view, where we can 
consider the act of picking one point from our data set at random as distribution,
known as the _empirical_ distribution associated with that data set.

### Other Loss Functions

If we continue the above analysis for other losses, we might come across some
other well-known statistical quantities. Suppose that, instead of being judged
on _squared error_, we were judged on absolute error. Finding the minimizer
here is a bit more mathematically complicated, so let's explore this problem
numerically instead. 

Let's assume our data comes from an _exponential_ distribution with mean $3$.
We can create samples from this distribution using the `rexp()` function in `R`:

```{r}
## Generate some exponential random variables
## Here rate=1/3 corresponds to mean = 3
X <- rexp(20, rate = 1/3)
```

We can define a _function_ that computes: 

$$f(a) \E[|X - a|]$$

for any guess $a$. To do this, we'll generate a _large_ number of $X$'s and 
use the fact that the _sample_ expectation will converge to the _population_
expectation (Law of Large Numbers). 

```{r}
abs_error_expected_loss_exponential <- function(a, n=10000){
    X <- rexp(n, rate=1/3)
    # I have to be a bit fancy here to get around the fact that R's
    # mean() will only return a single value.
    # If you want something simpler, but a bit slower, use the map() function
    rowMeans(abs(outer(a, X, `-`)))
}
```

Now we can try out a bunch of different values of $a$ and see what the best 
guess is: 

```{r}
x_guess <- seq(0, 6, length.out=501)
x_guess_loss <- abs_error_expected_loss_exponential(x_guess)

plot(x_guess, x_guess_loss, type='l',
     xlab=expression(a), 
     ylab=expression(E[X * "~ Exp(3)"] (abs(X - a))))
```

If we look at this curve more closely, we see that the best guess _for this
loss_ is not $3$, even though $3$ is the mean. Choosing a different loss 
function clearly changes the best guess. But what is that best guess?

If we look closely, we can see that the minimizer is right around 
`r x_guess[which.min(x_guess_loss)]`. But what is that value? 
After a bit of browsing the 
[Wikipedia page for the exponential distribution](https://en.wikipedia.org/wiki/Exponential_distribution), we
can see that it is suspiciously close to the _median_ of the distribution, which
here is $3 \log(2) \approx 2.08$. 

So our guess here is that the _median_ is the best possible guess. To confirm
our hunch, let's try out another distribution. The _log-normal_ distribution
is skewed, so its mean and median are quite different. Specifically, let's take

$$X \sim \text{LogNormal}(3, 2^2) \Leftrightarrow \log X \sim \text{Normal}(3, 2^2)$$

Reviewing the [standard formulas](https://en.wikipedia.org/wiki/Log-normal_distribution), 
we see that this distribution will have mean $e^{3 + 2^2/2} = e^7 \approx 1097$
but median $e^3 \approx 20$. This is a lot of skew and will let us clearly
see how much the median beats the mean. 

```{r}
abs_error_expected_loss_logn <- function(a, n=50000){
    X <- rlnorm(n, meanlog=3, sdlog=2)
    # I have to be a bit fancy here to get around the fact that R's
    # mean() will only return a single value.
    # If you want something simpler, but a bit slower, use the map() function
    rowMeans(abs(outer(a, X, `-`)))
}

# Need a nice big range here, but we need log-spaced values as well
x_guess <- 10^seq(log10(1), log10(1500), length.out=5001) 
x_guess_loss <- abs_error_expected_loss_logn(x_guess)

plot(x_guess, x_guess_loss, type='l',
     xlab=expression(a), log="xy",
     ylab=expression(E[X * "~ LogN(3, 2^2)"] (abs(X - a))))

abline(v=exp(3), col="blue4")
abline(v=exp(7), col="red4")
```

Here the blue line is the median and the red is the mean. You'll note it's not
perfect - the log-normal is a pretty 'heavy-tailed' distribution so these things
might take _many_ more samples to converge - but hopefully you can at least see
that the mean (red) is definitely not the best point. 

::: {.callout-tip title="What About 'Exact' Answers?"}

Suppose that instead you were graded on an 'exact match' loss: that is, 

$$L(\hat{y}, y) = \begin{cases} 1 & \hat{y} = y \\ 0 & \hat{y} \neq y \end{cases}$$

What would the optimal guess be now? 

*Hint*: It's another 'm' data summary. 

Note that this loss function is pretty unnatural for continuous values, but
is perfectly reasonable for classification. 

:::

### Custom Loss Functions

While the mean and median are well-understood statistically, it's not clear
that they are the _right_ answer for most problems. For how many things in life
is a _quadratic_ (squared error) loss exactly correct? 

Often, some additional predictive power can be attained for a given model by
determining a _bespoke_ loss function for a problem. For example, one of the
most confusing parts of the US tax code is that one has to guess how much one
will owe in taxes at the end of the year before you compute your taxes: estimate
too much and you're giving the government and interest-free loan (and missing
out on that sweet compounding interest), estimate too little and the IRS will
charge you interest on the money you failed to pay. And, of course, the IRS
interest that you get charged is often _much_ higher than what your bank would
pay to you. 

In this case, we have an _asymmetric_ outcome and our loss is something like: 

$$L(\hat{y}, y) = \begin{cases} 0.02|\hat{y} - y| & \hat{y} > y \text{ (Overpaid estimated taxes)} \\
0.07|y - \hat{y}| & \hat{y} \leq y \text{ (Underpaid estimated taxes)} \end{cases}$$

What should we do here? It's not as obvious, but hopefully you can convince
yourself that we definitely want to pay more than our _median_ guess since
things are asymmetric. (In fact, this type of loss moves away from the median
to more general quantiles.)

This family of losses is known as a check loss for (hopefully) obvious reasons:

```{r}
irs_loss <- function(delta_y) {
    ifelse(delta_y > 0, 0.02 * abs(delta_y), 0.07 * abs(delta_y))
}

d <- seq(-3, 3, length.out=201)
plot(d, irs_loss(d), xlab=expression(hat(y) - y), ylab="Financial Loss", 
     main="Check Loss", type="l")
```

In classification problems, this asymmetry is often even more stark: when
designing a medical test for cancer, there are two bad outcomes: 'false positive'
when you tell someone they might have cancer and send them for unnecessary
expensive follow-up tests and 'false negative' when you tell them they don't
have cancer and leave their tumors untreated and let them metastasize. Both bad,
but here 'false negative' is much worse and clinical guidelines should reflect
that. 

::: {.callout-tip title="Value of a Human Life"}

A key question in medical economics is how much is a _human life_ 
actually _worth_? While we would in a perfect world want everyone to receive
all possible care, in a finite world decisions have to be made about how to
ration finite medical resources. (In the US, we do this through
infinitely-complicated health insurance markets; in more centralized systems
like the UK, this decision is made by NHS analysts and economists.)
As societies become richer, they can generally put a higher dollar value on
human life. 

What do you expect this to do to the false negative/false positive
balance associated with cancer tests? Why?

:::

It's worth re-emphasizing: means and medians are the _exact_ right answer for
certain loss functions and we're not claiming to beat (conditional) means for
MSE loss; that would be impossible. Instead, by using a loss function that is 
more tailored to the actual human impact of a problem, we might get a better 
overall outcome. ("Better an approximate answer to the right question ..." and
all that.)

Building a custom loss function _can_ be effective for important problems, but
it requires some effort and often a bit of custom software. For more low-stakes
problems, the quadratic or absolute losses are 'good enough' to get started with
a problem.

### Metrical Determinism

In many ways, the choice of _loss_ function drives the behavior of ML systems - 
a phenomenon called _metrical determinism_ - and some of the most interesting
debates in ML relate to properly building a loss function. A good loss function
for numerical prediction or binary classification isn't that hard to come by, 
but what is the right performance measure for an agentic AI system? 

The news is currently full of debates around 'reward hacking' and agents. A few
years ago, an AI system was designed to play the classic video game Tetris. In
Tetris, you never win or beat the game; you just survive longer than any other
player. So when the AI system was set loose on Tetris, it tried a bunch of
strategies and quickly found that the optimal strategy had nothing to do with
`T` and `L` blocks, but was simply to hit `Pause`. As long as the game was 
paused, the player never lost and the survival time kept going up and up. 

In this case, the system clearly found an aspect of the loss function that didn't
quite align with actual human intent. Here, this was pretty easy to fix: either
disallow the pause button or change the loss function to capture 'blocks destroyed'
rather than 'time survived'. Either way, the problem arose because the original
loss function given to the machine didn't capture the whole set of human 
preferences implicit in the task. 

While the Tetris experience is cute, this behavior can be quite worrying at 
scale. To pick an example from the news: an agent is told that its goal is to do
well on some sort of numerical score as evaluated by a third-party website; the
agent can try to do well on the task, but isn't it easier to hack into the
'grader' and change the scores? 

Think through this sort of reward hacking for other problems: what happens if
an AI agent is asked to help minimize power consumption at a data center? 
SciFi is full of all sorts of takes on this fundamental problem. 

### Empirical Risk Minimization

Bringing things back to Earth, what does this imply about our task
building 'simple' predictive models? Our goal is simple, we want to find
the best predictor $f^*$ that minimizes our expected loss, *aka*, the 
'risk' of that predictor: 

$$R(f) = \E_{(\tilde{X}, \tilde{Y}) \sim \Dcal}[L(f(\tilde{X}), \tilde{Y})]$$

Conceptually, this should be a relatively simple problem. We just try an 
enormous number of $f$s and pick the one that minimizes $R(\cdot)$. This is
a bit impractical, but smart algorithms exist, so let's put the computational
concerns aside for a moment. 

Can we ever actually know $R(\cdot)$? Knowing $R(\cdot)$ exactly requires knowing
the full distribution $\Dcal$ relating $X$ and $Y$. If we really knew this, 
we wouldn't need ML in the first place. So now our task is to minimize a risk
we can't compute. 

But we have something else in our toolbox: our _training set_ of samples. We
don't know $R(f)$, but we can _approximate_ it with

$$R_n(f) = \frac{1}{n} \sum_{i=1}^n L(f(X_i), Y_i)$$

This is known as the _empirical_ risk since it captures the risk we observe
on our training set. It's not a perfect estimate of the true (population) risk,
but hopefully with enough data, it's good enough. Specifically, we hope
that the $f$ that minimizes $R_n(\cdot)$ is 'close enough' to the one that 
minimizes $R(\cdot)$. This can give us a 'nearly optimal' predictor, even if 
true optimality is beyond our grasp. This strategy is known as _empirical risk
minimization_ (ERM). 

To recap: ERM gives us a 'meta-formula' for finding a predictor. Given
some training data and a loss function, we pick the predictor that minimizes
the empirical risk, in hopes that it almost-minimizes the population risk. 
In symbols: 

- Let $f_*$ be the _true_ (Platonic) best predictor: 

  $$ f_* = \argmin_f R(f) = \argmin_f \E_{(\tilde{X}, \tilde{Y}) \sim \Dcal}[L(f(\tilde{X}), \tilde{Y})]$$
  
  We can't know $f_*$, but we hope to at least get close. 
- Let $\hat{f}$ be the _empirical_ (ERM) best predictor: 

  $$ \hat{f} = \argmin_f R_n(f) = \argmin_f \frac{1}{n} \sum_{i=1}^n L(f(X_i), Y_i)$$
- ML 'works' if these quantities are close: 

  $$R(\hat{f}) \approx R_n(\hat{f})$$
  
  The difference between these two quantities is known as the 
  _generalization gap_. When it is small, our model performs about as well on 
  the training set as it does on the test set. When the gap is large, past
  performance tells us nothing about future performance, so it's hard to trust
  a system to work well. 
  
*Notation Note*: Above, I used the notation $\hat{x} = \argmin_x h(x)$. This
means that $\hat{x}$ is the _minimizer_ of some function $h(\cdot)$. You will
also see $\hat{h} = \min_x h(x)$, where $\hat{h}$ is the _minimum_. In brief, 
$\argmin$ is the minimizing input to $h$ and $\min$ is the minimum output of 
$h$ and these will be related by $\hat{h} = h(\hat{x})$. For example, if 
$h(x) = (x - 3)^2 + 2$, we will have $3 = \argmin h$ and $2 = \min h$. Can you
see why?
  

## OLS within ML Taxonomy

### Linear Regression

Ok, with all that said, let's see how ERM plays out for our beloved OLS. 
Before we can do so, let's review how it fits into our taxonomy. Recall that 
linear regression is a method that makes predictions using a linear 
combination of inputs: 

$$\hat{Y} = \beta_0 + \beta_1 * X_1 + \beta_2 * X_2 + \dots + \beta_p X_p$$

Here, we assume we have $p$ inputs -- $X_1$ to $X_p$ -- each of which get their
own weights ('coefficients') $\beta_1$ to $\beta_p$. The 'inputs' are often called
'features', because they record some aspect of the instance we want to make a
prediction about, or 'covariates', because they (hopefully) co-vary (go up or
down together with) the response. 

We also assume that we have a constant _intercept_ term $\beta_0$ which gets 
added directly to all predictions, regardless of the specific values of 
$X_1, \dots, X_p$. The intercept is sometimes 'hidden' by instead creating
a 'constant feature' of all $1$'s and using that instead. 

$$\hat{Y} = \beta_0 X_0 + \beta_1 * X_1 + \beta_2 * X_2 + \dots + \beta_p X_p \text{ where } X_0 \text{ is always } 1$$

This is a bit more mathematically convenient - since $\beta_0$ is now not 
'special' in any way - but can be a bit confusing if you're not used to it. 
You should assume that linear regression essentially is always fit with an 
intercept term unless specifically stated otherwise. 

### Linear Regression Taxonomy

So how does linear regression fit into our taxonomy from last time?

Clearly, it is a _supervised_ method: we have a specific outcome, traditionally
called $Y$, that we want to predict. Additionally it is a _regression_ model
(as the name suggests), since our outcome is numeric, not categorical. 
Finally, OLS requires _numerical, tabular_ input data for the $X$s as well.
If some of the input features are categorical, we have to transform them into
numerical quantities before building a model; this process is known as 
_embedding_ and you will explore it more in teh first problem set. 

### OLS vs Linear Regression

So what is the difference between linear regression and ordinary least squares?
So far, we've set up an arbitrary linear model defined by some coefficients
$\bbeta$. But how do we ever pick which one? There are (infinitely!) many 
possible linear regressions. This is where our 'meta-rule' of ERM can provide
guidance. 

ERM tells us that we should pick the line that minimizes our loss. But what loss
does OLS use? The hint is in the name: least _squares_. OLS uses - and is defined 
by - MSE loss. 

So to fit OLS we pick the $\bbeta$ that minimizes our squared error loss _on
the training set_: 

$$\hat{f}_{\text{OLS}} = \argmin_f \frac{1}{n}\sum_{i=1}^n (Y_i - f(X_i))^2$$

Well - not quite! In the problem above, we actually don't have any guarantee
that the $\argmin$ will actually be a linear model. To ensure that, we add
a _constraint_: 

$$\hat{f}_{\text{OLS}} = \argmin_{f: f \text{ is a linear model}} \frac{1}{n}\sum_{i=1}^n (Y_i - f(X_i))^2$$

Here, we now saying that we're not picking the best $f$ of _all possible 
functions_; we're just picking the best $f$ of _all linear models_. So
"linear models" defines our acceptable _model class_ or _model family_ for
this problem. 

We've waited a little while to introduce it, but picking an acceptable
model family is _extremely_ important. (If you just pick the 'best possible
function' over all functions, solving any ERM is basically impossible.) The 
choice of model family is a huge point of leverage / control in building an ML
system. If we look at bigger / more complex families, we can do better, but
there's also more potential for things to go poorly; if we use smaller / simpler
families, the worst case is less bad, but the best case is less good. 

We often talk about model classes in terms of _size_, which is roughly the number
of parameters in the model. Linear regression has relatively few parameters (only
one per input), while giant LLMs are chosen from a very _large_ family
and have a huge number of parameters. 

Consider two model classes $\Fcal_1, \Fcal_2$ where $\Fcal_1$ is a
subset of $\Fcal_2$ ($\Fcal_1 \subset \Fcal_2$): that is, everything in $\Fcal_1$
is also in $\Fcal_2$, but some things in $\Fcal_2$ might not appear in $\Fcal_1$. 
You should be able to convince yourself of the following: 

$$\min_{f \in \Fcal_2} R(f) \leq \min_{f \in \Fcal_1} R(f) \leq \max_{f \in \Fcal_1} R(f) \leq \max_{f \in \Fcal_2} R(f)$$

Why is this true? (*Hint:* Who is taller? The tallest person on Columbia's 
campus or the tallest person in all of NYC? How do you know this?)

This formalizes what we said before: "If we look at bigger / more complex 
families, we can do better, but there's also more potential for things to go 
poorly; if we use smaller / simpler families, the worst case is less bad, but 
the best case is less good."

Next week, we'll talk about factors that lead to us getting a good $\hat{f}$
*vs* a bad $\hat{f}$ and how we can hopefully keep the generalization gap small,
but first (next class), we'll look at OLS in practice and examine how things
like test error, training error, generalization gap, *etc* can be computed.




