---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}

Outline: 

--- 

Today we're going to look a bit more closely at the _reasons_ why an ML system
can fail. Before we do so, let's review our basic framework of ML: 

$$\begin{align*}
\hat{f} &= \argmin_{f \in \Fcal} \frac{1}{n}\sum_{i=1}^n L(f(\bX_i), Y_i) \tag{ERM} \\
f_* &= \argmin_{f \in \Fcal} \E[L(f(\bX), Y)] \tag{Optimal $f_*$ Minimizes Test Error} \\
\frac{1}{n}\sum_{i=1}^n L(f(\bX_i), Y_i) &\approx \E[L(f(\bX), Y)] \text{LLN}
\end{align*}$$

Reviewing our pieces: 

- In the supervised context, we measure success on a _loss_ function $L$, where
  smaller is better
- The best possible prediction function (within the category $\Fcal$), $f_*$, 
  is the one that minimizes our test error, *i.e.*, our error on _new_ data
- Because we can't actually ever know our true test error, we instead typically
  pick our guess $\hat{f}$ as the best predictor (within the category $\Fcal$)
  on our _training_ data, *i.e.*, the one with the smallest empirical (training)
  risk
- Hopefully, the empirical risk (training error) is a good proxy for the 
  population risk (test error) and $f_* \approx \hat{f}$
  
This story seems coherent, but there are lots of places it can break! 

- Is the _loss_ function $L$ a true measure of what we actually want to happen? 
  (Think of reward hacking / alignment issues, though these are less common in
  the purely numerical context.)
- Is $R_n(f)$ actually a good proxy for $R(f)$? Any time we are estimating 
  something from data, we have to worry about sampling problems: 
  - Is our training data _representative_ of the (future) population we actually
    care about or is it biased in some way?
  - Is our training sample _large enough_ that the LLN actually 'kicks in'?
    If $R_n(f)$ is super noisy, minimizing it might not actually get us very 
    close to the minimum of $R(f)$. 
- In the ERM context, picking a large $\Fcal$ gives us a lower minimizer. How
  big is big enough? 
  
  $$\Fcal_1 \subseteq \Fcal_2 \implies \min_{f \in \Fcal_2} \hat{R}_n(f) \leq \min_{f \in \Fcal_1} \hat{R}_n(f)$$ 
  
  and 
  
  $$\Fcal_1 \subseteq \Fcal_2 \implies \min_{f \in \Fcal_2} R(f) \leq \min_{f \in \Fcal_1} R(f)$$ 
  
  (Confirm for yourself why these two statements are true.)
- Does _risk minimization_ actually mean we're doing what we want? Under 
  what conditions does $R(f_1) \approx R(f_2)$ imply $f_1 \approx f_2$? 
- How well can we _actually_ find $\hat{f}$? If $L$ is complicated, our data
  set is large, or $\Fcal$ is large, solving the ERM problem might not be 
  easy. 
  
Our goal over the next few classes is to look at several of these problems in
turn. Some of them are things we can get a handle on mathematically, while others
require more qualitative / problem-specific analysis. 

## Bounding the Generalization Gap

Suppose we have the _right_ (best) predictor $f_*$ and our ERM-estimated predictor
$\hat{f}$. We know that

$$\hat{R}_n(\hat{f}) \leq \hat{R}_n(f_*)$$ 

and 

$$R(\hat{f}) \geq R(f_*)$$ 

The first is true by ERM - $\hat{f}$ minimizes the training error $\hat{R}_n$ -
while the second is true by definition - the best predictor $f_*$ is defined
as the minimizer of $R(\cdot)$. While this seems like a nice tidy story, this
actually isn't quite enough to give us what we want: 

$$ R(\hat{f}) \leq R(f_*) + \text{A Little Bit} $$

That is, we hope that _in the future_ / _on the test set_, $\hat{f}$ will be
_nearly optimal_. To ensure something like this holds, we need to start unpacking
the sources of error. Let's start with the simplest: 

- $\Fcal$ might not actually contain the _true_ relationship. For example
  $\Fcal$ might be the set of all lines (as in OLS), but the _real_ relationship
  might be quadratic. In this case, even if we get the _best possible_ line, 
  the misfit between $\Fcal$ and the real world is an unavoidable source of 
  error. Let's call $f_{**}$ the _true_ best possible predictor - from all of
  math, not just $\Fcal$. We then have
  
  $$R(\hat{f}) - R(f_{**}) = \left(R(\hat{f}) - R(f_*)\right) + \underbrace{\left(R(f_*) - R(f_{**})\right)}_{\text{Systematic / Approximation Error}}$$
  
  We can't say much about this systematic error yet, other than to note that 
  a bigger $\Fcal$ will hopefully get $f_*$ closer to $f_{**}$. 
  
- Turning to the first term, we have

  $$ R(\hat{f}) - R(f_*) = \underbrace{\left(R(\hat{f}) - \hat{R}_n(\hat{f}) \right)}_{V_1} + \underbrace{\left(\hat{R}_n(\hat{f}) - \hat{R}_n(f_*)\right)}_B + \underbrace{\left(\hat{R}_n(f_*) - R(f_*)\right)}_{V_2}$$
  
  We know that $B \leq 0$ - why? So we can change this to an inequality like
  
  $$ R(\hat{f}) - R(f_*) \leq \underbrace{\left(R(\hat{f}) - \hat{R}_n(\hat{f}) \right)}_{V_1}  + \underbrace{\left(\hat{R}_n(f_*) - R(f_*)\right)}_{V_2}$$
  
  But what can we do with these two $V$ terms? Well, if we look closely at them
  both, they measure 'how far can $\hat{R}_n$ be from $R$ for a given $f$'. It's
  quite hard to say how much the empirical risk of a specific function can vary
  from its mean (Why is $\E[\hat{R}_n] = R?$) but we can generally get an upper
  bound on this quantity, so let's just look at the _maximum_ it can be: 
  
  $$ \begin{align*}
  R(\hat{f}) - R(f_*) &\leq \underbrace{\left(R(\hat{f}) - \hat{R}_n(\hat{f}) \right)}_{V_1}  + \underbrace{\left(\hat{R}_n(f_*) - R(f_*)\right)}_{V_2} \\
  &\leq \left|R(\hat{f}) - \hat{R}_n(\hat{f}) \right|  + \left|\hat{R}_n(f_*) - R(f_*)\right| \\
  &\leq 2 \max_{f' \in \Fcal} \left|R(f') - \hat{R}_n(f') \right| 
  \end{align*}$$

So if we put this all together, we get

$$ \begin{align*}
R(\hat{f}) &\leq R(f_{**}) + \left|R(f_*) - R(f_{**})\right| + 2  \max_{f' \in \Fcal} \left|R(f') - \hat{R}_n(f') \right| \\
           &= R(f_{**}) + \min_{f \in \Fcal} \left|R(f) - R(f_{**})\right| + 2  \max_{f' \in \Fcal} \left|R(f') - \hat{R}_n(f') \right| \\
\end{align*}$$

This is a key relationship, so let's look very closely at it: 

- On the left hand side, we have the _test error_ we want to control. All
  of our eventual choices should be made to try to make this as small as possible,
  but we can't really know it exactly, so we will make do by making our upper
  bound (on the right hand side) as small as possible. 
- The first term on the right-hand side, $R(f_{**})$ is the risk associated with
  the Platonic ideal of prediction; this _irreducible_ error is just a fact of
  nature and so we just take it as is. Sometimes, we can't do well on a task 
  because it's just hard. (*E.g.*, predicting social behavior from genetics, 
  there's definitely a little signal there, but even the best possible genetic
  model can't account for all the social factors that make us who we are.)
- The second term on the right-hand side, $R(f_*) - R(f_{**})$, is a 
  _systematic_ or _approximation_ error term. Even if we got the very best
  element of $\Fcal$, it can't be a perfect fit. Here, _we_ are the source of
  the problem. If we could use a 'better' (bigger) $\Fcal$, our error would be
  smaller. But, as we will soon see, this isn't a painless choice. 
  
  Because this term measures 'misfit', it is also often called 'bias' 
  in statistical lingo. (This isn't usually bias in the sense of prejudice, but
  it can be in some problems; it's closer to _bias_ in the sense of being 
  a little off center / slanted that you might hear in engineering or cooking.)
- The third term on the right hand side, 
  $\max_{f' \in \Fcal} \left|R(f') - \hat{R}_n(f') \right|$, is a measure of 
  _sampling_ or _estimation_ variability and it attempts to capture how far
  our _estimate_ of performance will be from true performance. If this is small,
  then picking $\hat{f}$ with good training performance should guarantee good
  out-of-sample performance. But if it is large, we don't have any such guarantee. 
  
  Note that sampling variability is not bad _per se_: we could (theoretically) 
  hope to do _better_ than we saw in the training data, but since we picked 
  $\hat{f}$ to be optimal on the training data, it seems pretty unlikely that 
  we'll beat it 'by chance.'
  
How then can we control

$$ 2\max_{f' \in \Fcal} \left|R(f') - \hat{R}_n(f') \right|$$

? Obviously, in a perfect world, we would just have lots of samples fo
$\hat{R}_n \approx R$ for all $f$ (LLN) and we'd be happy. In practice, we
don't just have a 'give me more data' option, so we have to look elsewhere. 
If $\Fcal$ were smaller, this quantity would be smaller; and larger $\Fcal$
will *ceteris paribus* give us a larger value. So minimizing this term makes
us want to pick as small an $\Fcal$ as possible. 

So here we have our dilemma: we want $\Fcal$ to be as large as possible to make
the approximation error small, but at the same tiem we want $\Fcal$ to be
as small as possible to make the sampling error small. Sadly, we can't have
both... More data is the only 'free win', but that's typically easier said
than done. 

### Aside: Is $R(f_{**})$ real? 

You might ask whether $R(f_{**})$ is actually real or whether it's just a
different type of approximation error. This is a bit philosophical (does true
randomness exist or do we just need ever more complex models?) so we won't get
into it here. It's not a term that will drive our decision making, so arguing 
whether it is exactly 0 or some other unknown, unknowable, and uncontrollable
number doesn't matter too much at the end of the day. 

## Visual Representation

Draw the 'two circles' picture here. 

Key point: $$\V(\hat{f}) \lesssim \frac{1}{2} |\mathcal{F}|$$

Argue by analogy with a simple random variable $Z \in [a, b]$: $\V(Z) \leq \frac{1}{2}(b-a)^2$. 

Q: What is variance for non-numeric quantities? Consistency / repeatability. 

## Exact Decomposition

The above decomposition works for _any_ loss function, but it turns out the
math is especially nice if we use an MSE loss. 

Specifically, let's look at several quantities here: 

- $Y$: the _true_ label we want to predict
- $\E[Y|X]$: the _optimal_ predictor under MSE loss ($f_{**}$ above)
- $f_*$: the best $\mathcal{F}$-class predictor
- $\hat{f}$: the element of $\mathcal{F}$ we wind up using. 

In this case, we get the exact relationship: 

$$\begin{align*}
\E[(Y - \hat{f}(X))^2|X] &= \E[(Y - \E[Y|X])^2|X] + \E[(\E[Y|X] - f_*(X))^2|X] + \E[( f_*(X) - \hat{f}(X))^2|X] \\
&= \sigma_Y^2 + \text{Bias}^2 + \text{Variance}
\end{align*}$$

So here our terms are perfectly additive in a "Pythagorean" kind of way. We see
that we have three terms: 

- Irreducible Error: doesn't depend on anything we have any control over
- Approximation Error / Bias: depends only on our choice of $\mathcal{F}$ (which
  determines $f_*$)
- Sampling Error / Variance: driven the the randomness in our data leading to 
  different $\hat{f}$ on different data sets. 
  
Other loss functions will follow this relationship _approximately_ and it's good
for your intuition, even if it's only _exactly_ true (as equality, rather than
some inequalities) for MSE. 



## Tradeoff? 

The above is a _bias-variance_ decomposition. It is a fundamental law of how we
do the accounting for these types of problems. We haven't 'created or destroyed'
error, we've simply put it into buckets that we hope to better understand. 

Later in the course, we will see that sometimes (but not always), we can adjust
the size of $\Fcal$ so that $\text{Bias}^2$ goes up by 4 while Variance goes
down by 9. This is a clear win! But we don't always have this option. 

One of the most open and active questions in modern ML theory is why sometimes
picking a bigger $\Fcal$ seems to help on both bias and variance (or at least 
not hurt). Answers to this are many--we might not actually be measuring things
correctly, this is 'worst case' theory that doesn't apply to real life, there
are some other factors at play that drive results here--and all true to one 
degree or another. 

For us, in the simple scalar supervised context: it's sufficient to say that we
are conservative 'worst case'-minded in theory, but to hope that theory will 
still provide some practical advice for real problems. 

Finally, note that essentially none of these quantities can be known _in practice_. 
Next time, we will look at practical approaches for actually measuring performance.


## A Simulation Study

While this type of decomposition is something we cannot work
out _in applications_ because we don't know the underlying
way the world works, but we can see how this works out in a
simulation study. 

Here, we're going to assume that the world works as: 

$$\begin{align*}
X &\sim \mathcal{U}([-3, 3]) \\
Y|X &\sim \mathcal{N}(3 + 2x + x^2, 2^2)
\end{align*}$$

and we will try to fit a line to this non-linear relationship. 

A set of $n=40$ points from this model might look like this: 

```{r}
#| echo: false
library(tidyverse)

set.seed(3106)

make_xy <- function(n){
    X <- runif(n, -3, 3)
    Y <- rnorm(n, 3 + 2 * X + X^2, 2)
    data.frame(X=X, Y=Y)
}

TRAINING_DATA <- make_xy(40) 

ggplot(TRAINING_DATA, aes(x=X, y=Y)) + 
    geom_point() + 
    xlab("X") + 
    ylab("Y") + 
    theme_bw() + 
    ggtitle("Example Training Data")
```


Given just this set of observations, it might not actually even be clear that
we 'should' be fitting a quadratic model. If we were to fit OLS to this data,
we'd see the following: 

```{r}
ggplot(TRAINING_DATA, aes(x=X, y=Y)) + 
    geom_point() + 
    xlab("X") + 
    ylab("Y") + 
    theme_bw() + 
    ggtitle("Example Training Data with Linear Fit") + 
    stat_smooth(se=FALSE, method="lm", color="red4")
```

Again - not obviously terrible!

So how good is this in practice? Let's start by computing the in-sample (training)
error: 

```{r}
#| message: false
library(broom)
m <- lm(Y ~ X, data=TRAINING_DATA)

augment(m) |> # Return original data along with `.fitted` (in-sample predictions)
    summarize(mse = mean((Y - .fitted)^2), 
              rmse = sqrt(mse))
```

So this is interesting: even using the 'wrong' model, we see that we have a
respectable RMSE. A bit worse than the 'true' of 2 from Nature's model, but not
terrible. 

How would we have done _out-of-sample_ instead? Let's create some test data from
Nature's model and compare: 

```{r}
TEST_DATA <- make_xy(n=1e6)

augment(m, newdata=TEST_DATA) |>
    summarize(mse = mean((Y - .fitted)^2), 
              rmse = sqrt(mse))
```

Ok - so definitely worse, which is what we would expect. 

Now that we've done a bit of this in practice, let's work out the parts of our
decompositions. We first need to work out $f_*$ - the *best* line we could have
fit to this data. Rather than figure out $f_*$ analytically, let's just see what
line we fit to an _enormous_ data set, taking advantage of the LLM so that
optimizing the fit to the huge data set is essentially fitting to Nature's model. 

```{r}
m_star <- lm(Y ~ X, data=make_xy(n=1e7))
m_star
```

So the 'best' line we could fit to this data is:[^six]

[^six]: If you're interested in where the $6$ comes from: work through 
$\E[3 + X^2]$ for $X \sim \mathcal{U}([-3, 3])$. 

$$f_*: \hat{Y} = 2X + 6$$

Visually, this looks something like: 

```{r}
X <- seq(-3, 3, length.out=201)
EY <- 3 + 2 * X + X^2
F_STAR <- 2 * X + 6
data.frame(X=X, Y=EY, F_STAR=F_STAR) |>
    ggplot(aes(x=X)) + 
    geom_point(aes(y=Y), color="blue4") + 
    geom_point(aes(y=F_STAR), color="red4") + 
    xlab("X") + 
    ylab("Y") + 
    ggtitle("Fitting the Optimal Linear Approximator (Red) to Non-Linear Reality (Blue)") + 
    theme_bw()
```

So what would the _test_ error associated with $f_*$ be: 

```{r}
make_xy(n=1e7) |>
    mutate(f_star = 2 * X + 6) |>
    summarize(mse = mean((Y - f_star)^2), 
              rmse = sqrt(mse))
```

A bit better than our estimated model above. At this point, we have some of our
terms sorted out: 

- Irreducible Error: $2^2$. There was nothing to do here _in silico_; we just 
  know it because we know how the data was generated. 
- Approximation Error: Because: 

  $$R(f_*) = \text{Irreducible Error} + \text{Bias}^2$$
  
  we have
  
  $$\begin{align*}
  \text{Bias}^2 &= R(f_*) - \text{Irreducible Error} \\
                &= 11.2 - 4 \\
                &= 7.2
  \end{align*}$$
  
  Alternatively, if we 'turn off' the noise in $Y$, we can just work this
  out analytically as $\E[(\E[Y] - f_*)^2]$: 
  
```{r}
X <- seq(-3, 3, length.out=401)
EY    <- 3 + 2 * X + X^2
FSTAR <- 6 + 2 * X

APPROXIMATION_ERROR <- mean((EY - FSTAR)^2)
APPROXIMATION_ERROR
```

  Note that this doesn't exactly match our simulation since there was a bit of
  randomness, but it's close enough. 
  
Finally, let's look into the _estimation_ error. This comes down to the fact
that our estimated model: 

```{r}
m
```

doesn't match the $f_*$ (infinite data) model: 

```{r}
m_star
```

Visually: 

```{r}
TRAINING_DATA |> 
    mutate(f_hat  = predict(m, newdata=data.frame(X)), 
           f_star = predict(m_star, newdata=data.frame(X))) |>
    ggplot(aes(x=X)) + 
    geom_point(aes(y=Y)) + 
    geom_line(aes(y=f_hat), color="green4") + 
    geom_line(aes(y=f_star), color="red4") + 
    theme_bw() + 
    xlab("X") + 
    ylab("Y") + 
    ggtitle("Comparison of Ideal (Red) and Fitted (Green) Linear Models")
```
From this photo alone, it's not actually clear which line
is better, so let's compute the errors on our training
set once again to remind ourselves: 

```{r}
# Training Error - Fitted Model
augment(m) |> 
    summarize(mse = mean((Y - .fitted)^2), 
              rmse = sqrt(mse))
```

and

```{r}
# Training Error - Fitted Model
augment(m_star, newdata=TRAINING_DATA) |> 
    summarize(mse = mean((Y - .fitted)^2), 
              rmse = sqrt(mse))
```

(Which of these is lower? Why?)

From the error we computed above, we can back out the estimation error for
this particular fit

```{r}
TOTAL_ERROR <- augment(m, newdata=TEST_DATA) |>
    summarize(mse = mean((Y - .fitted)^2)) |> 
    pull(mse)

IRREDUCIBLE_ERROR <- 4

ESTIMATION_ERROR <- TOTAL_ERROR - IRREDUCIBLE_ERROR - APPROXIMATION_ERROR
print(ESTIMATION_ERROR)
```

But this isn't the _variance_ per se. To really get a sense
of the variance, we need to see how much this quantity would change if we
had _different_ training sets. Let's code this up: 

```{r}
X <- seq(-3, 3, length.out=201)
EY <- 3 + 2 * X + X^2
F_STAR <- 2 * X + 6
G <- data.frame(X=X, Y=EY, F_STAR=F_STAR) |>
    ggplot(aes(x=X)) + 
    geom_point(aes(y=Y), color="blue4") + 
    geom_point(aes(y=F_STAR), color="red4") + 
    xlab("X") + 
    ylab("Y") + 
    ggtitle("Variance of OLS Estimation Around the Optimal Approximator (Red) to Non-Linear Reality (Blue)") + 
    theme_bw()

my_palette <- palette.colors(n=8)[-1]

for(i in 1:7){
    resample_training <- make_xy(n=40)
    m_resample <- lm(Y ~ X, data=resample_training)
    
    # We have to be a bit manual here with data= and mapping= since we are
    # plotting multiple training sets on a single figure
    G <- G + geom_point(data=augment(m_resample), 
                        mapping=aes(x=X, y=Y), 
                        color=my_palette[i], 
                        alpha=0.2) + 
             stat_smooth(data=augment(m_resample),
                         mapping=aes(x=X, y=Y), 
                         method="lm", 
                         se=FALSE, 
                         color=my_palette[i], 
                         alpha=0.2)
}

G
```

The thinner lines here are different realizations of the OLS
estimator if we had different training sets; in this case, we
definitely see some variability, but not too much as OLS is a 
rather stable method. 

TODO: Extend to polynomial regression. 
