---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}

Outline: 

--- 

## Challenges of High-Dimensional Regression

### Ridge Regression - Core Idea

## Constraints

### Measuring the Size of Coefficients

### Constraints and Penalties

Above, we write ridge regression as: 

$$\hat{\bbeta} = \argmin_{\bbeta: \|\bbeta\|_2 \leq \tau} \frac{1}{n}\sum_{i=1}^n L(\bbeta^{\top}\bx_i, y_i) = \argmin_{\bbeta: \|\bbeta\|_2 \leq \tau} \frac{1}{2}\|\by - \bX\bbeta\|_2^2$$

which is essentially the classical OLS problem (MSE ERM with a linear
predictor) with the additional constraint: $\|\bbeta\|_2 \leq \tau$ or,
slightly more conveniently, $\|\bbeta\|_2^2 \leq \tau^2$.[^sq] Here,
we're putting the 'ridge-ness' in the _constraint_ by requiring $\bbeta$ to not
be 'too big.'

This is conceptually elegant, but it does pose a few practical challenges: 

- How can we pick $\tau$? 
- How can we solve the above problem? In particular, what happens if our
  gradient points us 'beyond the bounds' of our constraint? 
  
These are definitely solvable (and indeed solved!) problems, but it certainly
would be nice if we could get rid of the constraint. And as it turns out, 
we can. 



[^sq]: Any norm is non-negative, so both terms of this inequality are
non-negative and squaring doesn't change the order.

## Ridge Regression


**These notes are from a previous class and are a bit beyond the level
of this course.**

In the [previous set of notes](./notes06.qmd), we saw that
$(\overline{X}_n)_{[-\beta, \beta]}$ could outperform the naive estimator
$\overline{X}_n$ if the true parameter is 'not too big'. 

Can we extend this idea to regression? 

Recall that we formulated OLS as: 

$$\hat{\bbeta}_{\text{OLS}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\|$$

We can apply a 'clamp' as: 

$$\hat{\bbeta}_{\tau-\text{Clamped}} = \argmin_{\bbeta \in \R: \|\bbeta\| < \tau} \frac{1}{2}\|\by - \bX\bbeta\| = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\| \leq {\tau}$$

Here, we have some choice in measuring the 'size' of $\hat{\bbeta}$: in fact, we
can *theoretically* use any of our $\ell_p$-norms. As with the squared loss, 
it turns out to be mathematically easiest to start with the $\ell_2$ (Euclidean)
norm. This gives us the so-called *ridge regression* problem: 

$$\hat{\bbeta}_{\text{Ridge}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2 \leq {\tau} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2^2 \leq {\tau}^2$$

Before we discuss solving this, let's confirm it is convex. Recall that for a
problem to be convex, it needs to have a convex *objective* and a convex *feasible
set*. Clearly the objective here is convex - we still have the MSE loss from OLS
unchanged. So now we need to convince ourselves that $\|\bbeta\|_2^2 \leq \tau^2$
defines a convex set.  We can look at this in 2D first: 

$$\|\bbeta\|_2^2 \leq \tau^2 \implies \beta_1^2 + \beta_2^2 \leq \tau$$

But this is just the equation defining a circle and its interior (in math speak
a 2D 'ball') so it has to be convex!

In fact, constraints of the form 

$$\|\bbeta\|_p \leq \tau \Leftrightarrow \|\bbeta\|_p^2 \leq \tau^p$$

are convex for all $\ell_p$ norms ($p \geq 1$). We will use this fact many times
in this course. The sets that this constraint defines are called $\ell_p$ 'balls'
by analogy with the $\ell_2$ figure, even if they aren't actually round.

```{r}
#| echo: false
#| fig.asp: 0.3
library(patchwork)
plot_ball <- function(p){
    title <- bquote(l[.(p)]~"Unit Ball")
    expand_grid(x=seq(-2, 2, length.out=201), 
                y=seq(-2, 2, length.out=201)) |>
        mutate(interior = abs(x)^p + abs(y)^p <= 1) |>
        filter(interior) |>
        ggplot(aes(x=x, y=y)) + 
        geom_point() + 
        coord_fixed() + 
        ggtitle(title) + 
        theme_bw() + 
        xlab("") + ylab("")
}
plot_ball(1) + plot_ball(2) + plot_ball(Inf)
```

So our ridge problem 

$$\hat{\bbeta}_{\text{Ridge}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2^2 \leq {\tau}^2$$
is indeed convex. 

While we can solve this without too much trouble (though we need to adjust
our basic gradient descent method to avoid going 'out of bounds'), it turns
out to be easier to use an optimization trick known as *Lagrange Multipliers* 
to change this to something easier to solve. 

The [Method of Lagrange Multipliers](https://en.wikipedia.org/wiki/Lagrange_multiplier)
lets us turn *constrained* optimization problems that look like: 

$$\argmin_x f(x) \text{ such that } g(x) \leq \tau$$

into *unconstrained* problems like: 

$$\argmin_x f(x) + \lambda g(x)$$

Here, the constant $\lambda$ is known as the Lagrange multiplier. Instead of 
_forcing_ $g(x)$ to be bounded, we simply _penalize_ any $x$ that makes $g(\cdot)$
large. For large enough penalties (large $\lambda$), we will eventually force
$g(x)$ to be small enough that we satisfy the original constraint. In fact, 
there is a one-to-one relationship between $\tau$ and $\lambda$, but it's usually
not possible to work it out in any meaningful useful manner; all we can say is that
larger $\lambda$ correspond to smaller $\tau$. 

::: {.callout-note title="Pause to Reflect" collapse="true"}

Why is this true? Why does small $\tau$ imply large $\lambda$ and *vice versa*?

:::

In this form, it's straightforward to pose ridge regression as: 

$$\argmin_{\beta} \frac{1}{2}\|\by - \bX\bbeta\|_2^2 + \frac{\lambda}{2}\|\bbeta\|_2^2$$

::: {.callout-note title="Pause to Reflect" collapse="true"}

Why were we allowed to put an extra $\frac{1}{2}$ in front of the penalty term?

:::

Since both terms of this expression are differentiable, it is easy to show 
that the solution to this problem is given by: 

$$\hat{\bbeta}_{\lambda} = (\bX^{\top}\bX + \lambda \bI)^{-1}\bX^{\top}\by$$

To wit: 

$$\begin{align*}
\bzero &= \frac{\partial}{\partial \bbeta}\left(\frac{1}{2}\|\by - \bX\bbeta\|_2^2 + \frac{\lambda}{2}\|\bbeta\|_2^2\right) \\
       &= -\bX^{\top}(\by - \bX\bbeta) + \lambda \bbeta \\
\bX^{\top}\by &= (\bX^{\top}\bX + \lambda \bI)\bbeta \\
\implies \hat{\beta}_{\text{Ridge}} &= (\bX^{\top}\bX + \lambda \bI)^{-1}\bX^{\top}\by
\end{align*}$$

There are several important things to note about this expression: 

- There are actually *many* ridge regression solutions, one for each value of
  $\lambda$. It is a bit improper to refer to "the" ridge regression solution. 
  We will discuss selecting $\lambda$ below. 
- If you are a bit sloppy, this looks something like: 

  $$\hat{\beta} \approx \frac{SS_{XY}}{SS_{XX} + \lambda}$$
  
  so we 'shrink' the standard OLS solution towards zero, by an amount depending 
  on $\lambda$. Higher $\lambda$ 
- The penalized form (as opposed to the constraint form) allows for $\hat{\beta}$
  to be arbitrarily large, **if the data supports it.**
  
  More than that, $\lambda$ gives us an explicit *trade-off*. When comparing
  two potential solutions ($\hat{\bbeta}_1, \hat{\bbeta}_2$), we will prefer
  $\hat{\bbeta}_1$ if: 
  
  $$\begin{align*}
  \frac{1}{2}\|\by - \bX\bbeta_1\|_2^2 + \frac{\lambda}{2}\|\bbeta_1\|_2^2 & \leq \frac{1}{2}\|\by - \bX\bbeta_2\|_2^2 + \frac{\lambda}{2}\|\bbeta_2\|_2^2 \\
  \|\by - \bX\bbeta_1\|_2^2 - \|\by - \bX\bbeta_2\|_2^2 & \leq \lambda(\|\bbeta_2\|_2^2 - \|\bbeta_1\|_2^2) \\
  \Delta \MSE &\leq \lambda \Delta \text{$\ell_2^2$-Norm}
  \end{align*}$$
  
  That is, we will take a solution with a MSE that is $\lambda$ lower if that lowers
  the squared-norm of the solution by 1. Here $\lambda$ measures the "price" we
  are willing to pay (in terms of MSE) for shrinking the norm of $\hat{\bbeta}$:
  at larger $\lambda$, we demand a *larger* MSE reduction *ceteribus paribus*
  to accept a larger $\hat{\bbeta}$, while at small $\lambda$, we are willing
  to take a larger $\hat{\bbeta}$ in exchange for a much smaller decrease in MSE.
  
- If we set $\lambda = 0$, we recover standard OLS. Continuing the 'price'
  discussion above, this is the limit in which we will take a larger $\hat{\bbeta}$
  for *any* decrease in training MSE. This makes sense as we originally built
  OLS as the ERM for MSE, by definition caring about *only* MSE reduction and
  nothing else. 
- Unlike the OLS solution, the ridge regression exists even when $p > n$ because
  the $+\lambda \bI$ term ensures the 'denominator' prefactor is always 
  invertible.
  
While this isn't as simple to analyze as some of the expressions we considered 
above, this looks similar enough to our 'shrunk' mean estimator that it's 
plausible it will improve on OLS. In fact, you can show that there is always 
_some_ value of $\lambda$ that guarantees improvement over OLS in expectation. 
(This is a result known as the "MSE Existence Theorem")

To wit, 

```{r}
#| warning: false
n <- 100
p <- 80
Z <- matrix(rnorm(n * p), nrow=n)
L <- chol(toeplitz(0.6^(1:p))) # Add 'AR(1)' correlation
X <- Z %*% L
beta <- runif(p, 2, 3)

eye <- function(n) diag(1, n, n)
calculate_ridge_mse <- function(lambda, nreps = 1000){
    MSE <- mean(replicate(nreps, {
        y <- X %*% beta + rnorm(n, sd=1)
        beta_hat <- solve(crossprod(X) + lambda * eye(p), crossprod(X, y))
        sum((beta - beta_hat)^2)
    }))
    
    data.frame(lambda=lambda, MSE=MSE)
}

lambda_grid <- 10^(seq(-2, 2, length.out=41))

RIDGE_MSE <- map(lambda_grid, calculate_ridge_mse) |> list_rbind()
OLS_MSE <- calculate_ridge_mse(0)$MSE # Ridge at lambda = 0 => OLS

ggplot(RIDGE_MSE, aes(x=lambda, y=MSE)) + 
    geom_point() + 
    geom_line() + 
    geom_abline(intercept=OLS_MSE, slope=0, lwd=2, lty=2) + 
    xlab(expression(lambda)) + 
    ylab("Estimation Error (MSE)") + 
    ggtitle("Ridge Regularization Improves on OLS") + 
    theme_bw() + 
    scale_x_log10()
```

Clearly, for smallish $\lambda$, we are actually outperforming OLS, sometimes
by a significant amount! For this problem, the minimum MSE (best estimate) actually
seems to occur near $\lambda \approx$ `r RIDGE_MSE |> filter(MSE == min(MSE)) |> pull(lambda) |> round(3)`
