---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}

Outline: 

- Model Selection & Sparsity
  - Why?
- Classical Methods
- Regularization Methods
- Lasso Demo
  - CV
  - Regularization Paths

--- 

Recall that last time we introduced the concept of _ridge regression_: 

$$\argmin_{\bbeta} \frac{1}{2n}\sum_{i=1}^n (y_i - \bx_i^{\top}\bbeta)^2 + \frac{\lambda}{2} \|\beta\|_2^2$$

(Remember: the extra $1/2$ factors here don't really make a difference when we're only interested in the _minimizer_ and they make doing
'calculus things' a bit cleaner: $\frac{d}{dx}\frac{x^2}{2} = x$.)

This was an _extension_ to our basic ERM principle: we don't just 
_match_ the training data as well as we can (OLS would dominate on that task)
but we do so while keeping our set of $\bbeta$ coefficients from getting too
big. Our goal is not 'small coefficients' _per se_, but small coefficients lead
to better generalizability (out of sample performance):

- Our model is less sensitive to inputs at both the fitting and prediction stage.
  - At fitting, we keep out values from getting too big, so the influence from
    any single observation is reduced. (Compare to 'normal' regression where one
    observation can have essentially unbounded influence / leverage.)
  - At prediction, if our coefficients are smaller, a change to one of the input
    $x_i$ gives a smaller change in the output $\hat{y}$
- Due to the statistical 'magic' of shrinkage, at least a little bit of ridge
  penalization always helps. (In practice, our theory only holds for 'infinitely
  small' shrinkage, but we find that an appropriately CV-tuned $\lambda$ is
  often not that small.)

We motivated this idea from the intuition that it is impossible for _all_ 
features to be super important (the chaotic piranha principle), but this wasn't
actually mathematically necessary. (I personally believe it is an important
aspect of ridge's practical successes.)

Let's take a closer look at what that ridge solution actually looks like. A 
little bit of matrix calculus (which you aren't required to reproduce) tells us
that the solution looks like: 

$$\hat{\bbeta}_{\lambda} = (\bX^{\top}\bX + \lambda \bI)^{-1}\bX^{\top}\by$$

Contrast this to 'normal' OLS which has the following matrix form: 


$$\hat{\bbeta}_{\lambda} = (\bX^{\top}\bX)^{-1}\bX^{\top}\by$$

If you recall your formulas for simple (univariate regression), the OLS slope is
given by something like $$SS_{XY}/SS_{XX}$$ and the matrix form above just
generalizes this to multiple features. Here $\bX^{\top}\bX$ plays the role of
$SS_{XX}$ and $\bX^{\top}\by$ and $\bX^{\top}\by$ plays the role of $SS_{XY}$. 
(We can't 'divide' by matrices, but it's fine to multiply by the inverse of
$\bX^{\top}\bX$; essentially, if we couldn't divide by 3, we can multiply by 
$1/3$ to get to the same end result.)

Extrapolating this analogy: the ridge solution looks something like: $SS_{XY}/(SS_{XX} + \lambda)$. That is, we grow the denominator by $\lambda$ and
effectively 'shrink' the OLS solution closer to zero.[^exact]

In this form, we see that the $\lambda \to 0$ limit gets us back to OLS, while
the $\lambda \to \infty$ limit pushes us all the way back down to zero. 

*Understanding Check*: How does this relate to the behavior of $\tau$ in the 
constraint form?

We see something else a bit magical about ridge regression here: because of the
$+\lambda$ term in the denominator, we avoid any division by zero problems. 
(Recall that $\lambda$ is a _price_ and is always positive.) 

This is actually pretty magical and it means we can fit ridge regression even in
the 'wide' data setting where $p > n$! Hence, even if we have more measurements
than samples, we can fit something 'OLS-like'. Of course, it isn't "true" OLS
since we have the ridge penalty, but if $\lambda$ isn't too big, it's basically
the same thing. 

This is _incredibly_ important in fields where getting additional measurements 
($p$) is cost effective, but getting additional samples ($n$) is expensive.
Think medical trials, consumer tracking, weather and climate monitoring (after you've bought and installed your weather monitor, letting it run for a longer 
period is cheap; buying a new one and paying someone to install it is expensive)
and many more. 

So what does a ridge regression solution actually _look like_ in that situation?

```{r wine_setup}
#| cache: true
# Generated with help of Gemini
library(readr)
library(dplyr)
library(stringr)
library(tidytext)
library(tidyr)
library(glmnet)

# 1. Ingest static TidyTuesday snapshot
raw_wine <- read_csv(
    "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-05-28/winemag-data-130k-v2.csv",
    show_col_types = FALSE
)

# 2. Subsample to create a controlled p > n setting (n = 100)
set.seed(9750)
wine_sample <- raw_wine |>
    filter(variety == "Pinot Noir", !is.na(points), !is.na(description)) |>
    slice_sample(n = 100) |>
    mutate(wine_id = row_number()) |>
    select(wine_id, points, description)

# 3. Tokenize, strip stopwords/digits, and prune singletons
tidy_words <- wine_sample |>
    unnest_tokens(word, description) |>
    anti_join(stop_words, by = "word") |>
    filter(!str_detect(word, "^[0-9]+$"))

frequent_words <- tidy_words |>
    distinct(wine_id, word) |>
    count(word) |>
    filter(n >= 2)

# 4. Reshape to wide DTM via pivot_wider()
dtm <- tidy_words |>
    semi_join(frequent_words, by = "word") |>
    count(wine_id, word) |>
    pivot_wider(
        names_from = word,
        values_from = n,
        values_fill = 0
    )

# 5. Build design matrix X and continuous target y
aligned <- wine_sample |>
    select(wine_id, points) |>
    inner_join(dtm, by = "wine_id")

X <- aligned |> select(-wine_id, -points) |> as.matrix()
y <- aligned$points

dim(X)
```

Here we have a data set with $n = 100$ rows corresponding to reviews of different
bottles of wine and $p > n$ words used within those reviews. Our outcome $Y$ is
the rating (on a 0-100 scale) given to each bottle. 

The `glmnet` package will let us fit ridge regression easily: 

```{r}
library(glmnet)
# alpha = 0 <=> ridge. Hold on for alpha = 1!
glmnet(X, y, lambda=1, alpha=0) |> 
    coef() 
```

Ok - so we get a lot of _nearly_ zero coefficients. What happens if we let 
$\lambda$ get bigger? 

```{r}
glmnet(X, y, lambda=100, alpha=0) |> 
    coef() 
```

You can make $\lambda$ even bigger, but you won't see many exact zeros. 

It turns out to be helpful to look at many $\lambda$ at once and smart algorithms
make this not too slow. The `glmnet` package can do this for us: 

```{r}
ridge_model <- glmnet(X, y, alpha=0)
plot(ridge_model)
```

Here, we have a so-called _regularization plot_. Each line is the coefficient
associated with one variable. As $\lambda \to \infty$, we see all the variables
get pushed towards zero, but never actually get there. Given our rough
$$SS_{XY}/(SS_{XX} + \lambda)$$ intition, this shouldn't surprise us. A function
like $3/(x + 2)$ gets very close to zero as $x$ grows, but we have to go 
'all the way to infinity' to _actually_ get an exact zero.  In fact, if you look
at the plot above, it has that same sort of 'asymptotic towards zero' behavior.
(Note that we're looking at $-\log \lambda$ so the axis might be flipped from
your intuition.)

Does this matter? If we're only interested in making good predictions, not 
really. The amount of shrinkage to get things all the way to zero would be
enough to screw up our predictive accuracy (over-shrinkage) and for many
circumstances it's totally fine to use all of our features. 

_Scientifically_, this can be a bit more problematic. We often use OLS not just
as a pure predictive black box, but as a tool to _understand_ what features
matter and don't matter in our data. Ridge regression using a little bit of
everything 'muddles' the picture and doesn't give us insight into what
really matters.[^in_defense]

Instead, we may want a model that _explicitly_ only uses a few features. 
Sometimes, this may be something we really believe, but more often, it's a sort
of 'simplification for the sake of interpretation.'

As an aside, `glmnet()` can also help us perform CV to pick a strong $\lambda$: 

```{r}
ridge_cv <- cv.glmnet(X, y, alpha=0)
plot(ridge_cv)
```

We'll talk more about the finer points of CV for ridge in a later class.

## Sparsity

Models which use only a small subset of the potential inputs are called _sparse_
models and you have likely used some version of them previously, if subtly. 
You may have fit a linear model and used some sort of iterative procedure to 
decide which features to add (forward stepwise?) or remove (backward stepwise?).
The final model you settled on was not obviously sparse, but you were 
effectively 'zero-ing out' all of the features you chose not to include.

When is this useful? 

- If you think that only a few features are genuinely relevant and most are
  truly zero ('true sparsity'), obviously fitting a sparse model makes sense.
  It's hard to think of too many situations where this is likely to actually
  be the case. 
- If you think that only a few features are _important_ and most are near zero,
  the practical benefits of zeroing out the 'near zeros' may be worth it. 
  These benefits might include: 
  
  - Interpretation
  - A bias-variance gambit
  - Measurement cost. If we're deploying an ML system, it may be expensive to
    measure _everything_ and if a model does 99% per as well at 1% of the 
    data collection cost, it might be worth it. 
  - Computational performance. In very high-performance settings, looking up
    and computing with many features is slower than only using a few. 
  
[^in_defense]: You might argue that we're looking at the wrong end of the plot
($\lambda \to \infty$) and that it's better to look at a more reasonable 
$\lambda$ where our predictions are accurate and to just pick the biggest 
coefficients there. This is an entirely defensible point-of-view. 


[^exact]: This is exactly true in the case where our features have no 
correlation. In practice, recall that the vast majority of 'big $p$' problems 
occur because we have a bulk measurement technology and most of those 
measurements are highly correlated. (Think of having temperature sensors all 
over a city or doing genetic processing on a patient's entire genome.)

How do we _quantify_ sparsity? Recall that last time, we used the $\ell_p$ norms
to measure bigness. Using a calculus-y limiting sort of argument, we can argue
the $\ell_0$ norm can be used to count the number of non-zero elements in a vector. 

```{r}
x <- c(1, 0, 2, 0, 3, 0, 4, 0, 5)

# Compute |x|_p^p. 
lp_norm_p <- Vectorize(function(x, p){ sum(abs(x)^p)}, "p")
lp_norm_p(x, 1) # As expected
lp_norm_p(x, 2) # Matches direct calculation sum((1:5)^2)
lp_norm_p(x, c(1, 0.5, 0.1, 0.05, 0.01))
```

As we can see, this is quickly converging to the value of 5, for 5 non-zero
elements. (You can play around with this example and see it holds generally.)

So we sometimes **define** $\|\bx\|_0$ as the number of non-zero entries in a
vector $\bx$. This isn't a _real_ norm ($\sqrt[0]{x}$ isn't really a thing...)
but it's convenient so we'll use this. 

In this set-up, many of the sparse regression methods you have seen in the past
can be written as: 

$$\argmin_{\bbeta} \frac{1}{n}\sum_{i=1}^n (y_i - \bx_i^{\top}\bbeta) + \lambda \|\bbeta\|_0 = \argmin_{\bbeta} \text{Training-MSE}(\bbeta) + \lambda \|\bbeta\|_0$$

Here, as with ridge regression, $\lambda$ is a 'cost' associated with $\beta$. 
But, in ridge regression, we paid a continuous 'cost' for making $\beta$ a bit 
bigger; here we only pay a cost for adding a new variable. Once it's included, 
we don't penalize the specific value. 

The interpretation of $\lambda$ here is maybe even more natural than in ridge
regression: it's the 'value add' a variable has to add to be included. 
Specifically,  if adding a variable reduces the MSE by more than $\lambda$ it
is worth it; otherwise, the variable gets left out. (Make sure you can see this!)

All the rules for model inclusion you have seen before: $p$-values, ANOVAs, 
$R^2$ differences, $F$-tests, *etc.* are just different rules for picking a good
$\lambda$. 

So that's awesome! As before, we can't just pick $\lambda$ to minimize training
error (or we'll push $\lambda \to 0$, include all variables, and recover OLS
our 'training-error maximalist'), so we need to use CV but otherwise we're good.
Right?

Sadly not quite. The above problem is lovely in theory, but actually quite hard
to compute in practice. That $\|\bx\|_0$ term is really 'jumpy' and doesn't play
nicely with our gradient descent algorithms. Applied math folks have fancy
algorithms to solve it, but they require expensive software and computers and
still fail around $p \approx 1000$, which isn't _that_ many features. (In our
wine example, where each word was a feature, it's easy to see how you could get
more than 1,000 features.)

So we cheat. Or, as the mathematicians would say, we 'relax the problem.' 
Instead of solving the optimal, but impossible, problem, we solve a closely
related, simpler problem, and hope that it's good enough.[^alt] Specifically,
what if we don't use $\ell_0$ but instead use some other 
$\ell_{p \approximately 0}$. Let's look at this graphically. 

Here, I show $|x|^p$ for different $p$. 

```{r}
#| message: false
library(tidyverse)
library(ggforce)
p_grid <- c(0, 0.1, 0.5, 0.9, 1, 1.5, 2, 4)
x_grid <- seq(-3, 3, length.out=201)

expand_grid(p=p_grid, x=x_grid) |>
    mutate(xp = abs(x)^p) |>
    ggplot(aes(x=x, 
               y=xp, 
               color=factor(p))) + 
    geom_line() + 
    theme_bw() + 
    scale_color_brewer(type="qual", palette=2, name=expression(p)) + 
    ylab(expression(abs(x)^p)) + 
    facet_zoom(ylim=c(0, 1)) + 
    geom_point(data=data.frame(x=0, xp=0, p=0), 
               size=2)
```

Notice how the $p=0$ path has a hard jump at 0 from $|0.0000001|^0 = 1$ but
$0^0=0$. (Not all software uses this rule; but remember we're using the zero
power to count non-zeros.)

[^alt]: When we can't give an _exact_ answer to the _exact_ problem we want to 
solve, we can either give an _approximate_ answer to the _exact_ problem or
an _exact_ answer to the _approximate_ problem. In this section, we're taking
the later route as it has been more successful and influential in ML; but there's
a long and proud history of giving a nearly-optimal _approximate_ solution 
as well. Neither universally dominates the other. 
