---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}

Outline: 

--- 

Today, we're going to take some of the _theoretical_ decompositions we
talked about last time and put them into practice. Recall that we argued-
as a matter of _definition_ and _accounting_-that test error could
be decomposed as something like: 

$$\text{Test Error} = \text{Irreducible Error} + \text{Structural Error} + \text{Estimation Error}$$

in general and, in the specific case of MSE, we can equate these to: 

$$\E[\text{TestMSE}] = \sigma_Y^2 + \underbrace{\E[(f_{**} - f_*)^2]}_{\text{Bias}^2} + \underbrace{\E[(f_* - \hat{f})^2]}_{\text{Variance}}$$

where $f_{**}$ is Nature's model, *i.e.*, the most perfect model any
mind could ever devise with unlimited computing power, data and time; 
$f_*$ is the _best_ model _within_ a certain class $\Fcal$, *e.g.*, the best 
possible linear model or the best decision tree; and $\hat{f}$ is the
model we actually wind up selecting (often in an ERM manner). 

This construction suggests that we face a balancing challenge in deciding what 
_model category_ to use: 

- Picking $\Fcal$ to be _large enough_ to make the gap between $f_*$ and 
  $f_{**}$ small; while also
- Picking $\Fcal$ to be _small enough_ to keep the estimation error under
  control. *Ceteris paribus*, a 'smaller' $\Fcal$ bounds the gap between
  $\hat{f}$ and $f_*$ more tightly and reduces our variance. 
  
Let's talk through some scenarios: 

- The 'true' relationship is linear and we are fitting a line 
  ($\Fcal = \text{Linear Models}$). Here, the approximation error is zero
  because $f_{**} \in \Fcal$ so _all_ the error we have a chance at handling
  is _variance_. This is the 'classical' version of OLS theory where concepts
  like 'BLUE' come into play. 
- The 'true' relationship is quadratic, but we are still fitting a line. 
  Here, there is some approximation error that can't be avoided, but the 
  variance is (basically) the same as before, so we should expect to do worse.
- Because we want to avoid approximation error, we now try a much larger set
  of models--$\Fcal = \text{All 5th Order Polynomials}$. Approximation error
  goes back to zero, but at the cost of greatly increased variance. Potentially,
  this scenario leads to worse error than 'just fit a line' because the variance
  cost is too high for the bias reduction. 
- We go back to fitting a line to a relationship that is actually non-linear. 
  But now we have _absolutely massive_ data ($n \approx \infty$). As the 
  data set gets larger, the variance goes to zero and we are left with only
  approximation error. In this case, even with 'infinite' data, our error doesn't
  ever go to zero. 
- Finally, now armed with our huge data set, we go back to the set of all 
  polynomials. Here, variance still goes to zero, but our model is sufficiently
  flexible that our approximation also goes to zero and we get 'effectively 
  optimal' predictions, leaving only the irreducible error. We can essentially
  get $\hat{f} = f_* = f_{**}$. 
  
This reveals the general value of _big data_ in ML. It gives us a source of
leverage to manage variance; we can then use our choice of $\Fcal$ to make
bias (near) zero. Put those together and we have great predictions. If we don't
have much data, the variance remains untamed and we can have arbitrarily 
terrible predictions. 

We see this joint scaling of model complexity and data throughout contemporary
ML practice. Big data centers are required not only to process big data sets
_per se_ but also to fit the big models that those big data sets allow. (The
exact form and rate of this scaling is a topic of significant economic interest:
if _doubling_ data set size and _4x-ing_ compute cost leads to a 2% increase
in accuracy, is it worth it? More generally, if performance grows _sub-linearly_
in cost, there is a point of diminishing economic return that could be hit.
Alternatively, if twice as much cost gives four times as much performance, it 
might make sense to scale in perpetuity.)

So let's think through other ways that $\Fcal$ can be 'bigger' or 'smaller'. 

- Fitting a model with non-linear terms (quadratics, cubics, trig functions, 
  exponentials) gives a bigger model class. 
- Fitting a model with _more_ features gives a bigger model class. The set of
  6 feature linear models is larger than the set of linear models that only uses
  the first feature. (To see this, note that a model with slope 2 and one feature
  is essentially the same as the six feature model with coefficients 
  $(2, 0, 0, 0, 0, 0)$.)
- When fitting a decision tree, a tree with _more levels_ is more complex than
  a smaller tree. 
- Bigger (more parameter) neural networks are larger _ceteris paribus_. 
- Any others? 

In each of these, note that $\Fcal$ is a choice that you, the modeler, make, 
not a feature of the data. Bigger data makes bigger $\Fcal$ a good choice, but
it is possible to fit use $\Fcal$ on a small-ish data set. (We just have
overfitting danger.)

In [PS#01](../problem_sets/ps01.qmd), you will take a look at $K$-nearest
neighbors. Let's take a look at how the $K$ in $K$-NN drives model complexity. 

For concreteness, let's suppose we have data generated from a rather crazy looking underlying process ($f_{**}$): 

$\E[Y|X] = f_{**}(X) = 2X + 3 + 4\sin(X^2) + 9\sqrt{\log(1 + X^4)}$

This gives a somewhat odd shape, but one where a linear fit would not be
necessarily awful: 

```{r}
library(tidyverse)
data.frame(X = seq(1, 5, length.out=501)) |>
    mutate(EY = 2 * X + 3 + 4 * sin(X^2) + 16 * sqrt(log(1 + X^4))) |>
    ggplot(aes(x=X, y=EY)) + geom_line() + 
    ggtitle(expression(f["**"])) + 
    xlab("X") + 
    ylab("E[Y|X]") + 
    theme_bw()
```

Of course, in practice, we do not really get to see this relationship and we
instead just get a noisy set of observations. 

```{r}
OBSERVED <- data.frame(X = runif(50, 1, 5)) |>
    mutate(EY = 2 * X + 3 + 4 * sin(X^2) + 16 * sqrt(log(1 + X^4)), 
           Y = EY + rnorm(50, sd=4)) |>
    select(-EY)
    
BASE_PLOT <- ggplot(OBSERVED, aes(x=X, y=Y)) + 
    xlab(expression(X)) + 
    ylab(expression(Y)) + 
    theme_bw() + 
    geom_point(alpha=0.5)

BASE_PLOT + ggtitle("Training Data")
```

Instead of fitting a line, let's instead break out $K$-NN. Recall that $K$-NN
works by taking the average of the $K$ 'nearest neighbors' to the test point, 
*i.e.*, the point at which we want to make a prediction. While finding nearest
neighbors in higher dimensions is a bit tricky (are all directions equally
important?), in 1D this is pretty straightforward: 

```{r}
#| message: false
library(tidymodels) 
TEST_X <- data.frame(X = seq(1, 5, length.out=501))

knn5 <- nearest_neighbor(neighbors = 5, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
  set_engine("kknn") |>
  set_mode("regression") |>
  fit(Y ~ X, data = OBSERVED)

Yhat_knn5 <- augment(knn5, new_data=TEST_X) 

BASE_PLOT + geom_point(aes(y=.pred), 
                       color="red4", 
                       data=Yhat_knn5) + 
    ggtitle("5-Nearest Neighbor Predictions")
```

A pretty reasonable fit. We see here that this model is 'jumpy' at points
where the set of nearest neighbors changes, and flat otherwise. (Why?)

Let's compare this to some other values: what happens if we set $K=1$? 

```{r}
knn1 <- nearest_neighbor(neighbors = 1, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
  set_engine("kknn") |>
  set_mode("regression") |>
  fit(Y ~ X, data = OBSERVED)

Yhat_knn1 <- augment(knn1, new_data=TEST_X) 

BASE_PLOT + geom_point(aes(y=.pred), 
                       color="blue4", 
                       data=Yhat_knn1) + 
    ggtitle("1-Nearest Neighbor Predictions")
```

Much more wiggly! This makes sense - by only 'averaging' one neighbor
(doing no averaging really), we aren't really taming the variance in the
data at all. 

How about if we make $K$ much larger, say 25?

```{r}
#| message: false
library(tidymodels) 
TEST_X <- data.frame(X = seq(1, 5, length.out=501))

knn5 <- nearest_neighbor(neighbors = 5, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
  set_engine("kknn") |>
  set_mode("regression") |>
  fit(Y ~ X, data = OBSERVED)

Yhat_knn5 <- augment(knn5, new_data=TEST_X) 

BASE_PLOT + geom_point(aes(y=.pred), 
                       color="red4", 
                       data=Yhat_knn5) + 
    ggtitle("5-Nearest Neighbor Predictions")
```

A pretty reasonable fit. We see here that this model is 'jumpy' at points
where the set of nearest neighbors changes, and flat otherwise. (Why?)

Let's compare this to some other values: what happens if we set $K=1$? 

```{r}
knn1 <- nearest_neighbor(neighbors = 25, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
  set_engine("kknn") |>
  set_mode("regression") |>
  fit(Y ~ X, data = OBSERVED)

Yhat_knn25 <- augment(knn1, new_data=TEST_X) 

BASE_PLOT + geom_point(aes(y=.pred), 
                       color="green4", 
                       data=Yhat_knn25) + 
    ggtitle("25-Nearest Neighbor Predictions")
```

Not surprisingly, this is much smoother since, at any point, we are averaging 25 points. (From basic statistical theory, we know that
our variance is 1/25th of what we had for $K=1$-NN.) But this is
probably a bit _too_ smooth. We've over averaged and are missing the
sinusoidal 'wiggles' that are present in $f_{**}$. 

How might we _pick_ $K$? We have talked a lot about ERM, so let's try it
here. We can compute the _training_ error at lots of different $K$s and
pick the one that works best. 

```{r}
make_training_set_predictions_knn <- function(k){
  knn_k <- nearest_neighbor(neighbors = k, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
  set_engine("kknn") |>
  set_mode("regression") |>
  fit(Y ~ X, data = OBSERVED)
  
  augment(knn_k, new_data=OBSERVED) |> 
      rename(Y_hat = .pred) |>
      summarize(training_rmse = sqrt(mean((Y - Y_hat)^2))) |>
      mutate(k=k)
}

map(1:25, make_training_set_predictions_knn) |> 
    bind_rows() |> 
    ggplot(aes(x=k, y=training_rmse)) + 
    geom_point() + 
    xlab(expression(K)) + 
    ylab("Training RMSE") + 
    theme_bw() + 
    ggtitle("Selecting the K in KNN via ERM - Training Error")
```

Hmmm.... This seems fishy. This seems to _really_ suggest suggest that
$K=1$ is ideal. But do we want that? And is $K=1$ giving zero error really
as great as it sounds?

Well - we can't normally do this - but we have the option to look at 
test error in simulation land. Let's see how it does. 

```{r}
BIG_TEST_SET <- data.frame(X = seq(1, 5, length.out=5001)) |>
    mutate(EY = 2 * X + 3 + 4 * sin(X^2) + 16 * sqrt(log(1 + X^4)), 
           Y = EY + rnorm(5001, sd=2))

make_test_set_predictions_knn <- function(k){
  knn_k <- nearest_neighbor(neighbors = k, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
  set_engine("kknn") |>
  set_mode("regression") |>
  fit(Y ~ X, data = OBSERVED)
  
  augment(knn_k, new_data=BIG_TEST_SET) |> 
      rename(Y_hat = .pred) |>
      summarize(test_rmse = sqrt(mean((Y - Y_hat)^2))) |>
      mutate(k=k)
}

map(1:25, make_test_set_predictions_knn) |> 
    bind_rows() |> 
    ggplot(aes(x=k, y=test_rmse)) + 
    geom_point() + 
    xlab(expression(K)) + 
    ylab("Test RMSE") + 
    theme_bw() + 
    ggtitle("Selecting the K in KNN via ERM - Test Error")
```

$K=1$ is a **terrible** choice here! ERM lead us astray. 

Why did this happen? The answer here is *overfitting* in its purest form.

$K=1$ gave us 0 error. Why did this happen? Why wouldn't this hold true
on the test set? These are questions you will explore in [PS#01](../problem_sets/ps01.qmd). 

## Hold Out Method

So from the above, we see that something around $K\approx 6$ seems optimal.
But we cheated here and can't generally just look at the test set. What
should we do instead? 

Well - we have to fake it so we can make it. If we don't / can't look at
a _true_ test set, we need to do the next best thing. 

We started with 40 data points and used them all to fit the model. Because
$K$-NN had all 40 points in its memory, our training error was a terrible
proxy for test error. Instead, we need to **hold-out** part of our data set
to test the model. 40 data points is a lot, so this will be painful, but it's
what we have to do. 

Let's set aside 10 points (25%) from our original data set to form a
_validation set_ and fit our model on the remaining 30 points as a 
reduced _test set_. 

```{r}
hold_out_ix <- sample(40, 10)

hold_out_set <- OBSERVED[hold_out_ix,]
training_set <- OBSERVED[-hold_out_ix,]

make_validation_set_predictions_knn <- function(k){
  knn_k <- nearest_neighbor(neighbors = k, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
  set_engine("kknn") |>
  set_mode("regression") |>
  fit(Y ~ X, data = training_set)
  
  augment(knn_k, new_data=hold_out_set) |> 
      rename(Y_hat = .pred) |>
      summarize(validation_rmse = sqrt(mean((Y - Y_hat)^2))) |>
      mutate(k=k)
}

map(1:25, make_validation_set_predictions_knn) |> 
    bind_rows() |> 
    ggplot(aes(x=k, y=validation_rmse)) + 
    geom_point() + 
    xlab(expression(K)) + 
    ylab("Test RMSE") + 
    theme_bw() + 
    ggtitle("Selecting the K in KNN via Validation Set Error")
```

This isn't great! But it is better than our 'pure ERM' approach. We will
probably pick a $K$ around the 'true' right answer of 6 and, if we compare
the $Y$-axes across our plots, our validation error at any $K$ is a more
reasonable approximation to the actual test error. Let's put them all on
a plot together to make this clearer: 

```{r}
reduce(list(
    map(1:25, make_test_set_predictions_knn) |>
        bind_rows() ,
    map(1:25, make_training_set_predictions_knn) |>
        bind_rows(),
    map(1:25, make_validation_set_predictions_knn) |>
        bind_rows()
), inner_join, join_by(k)) |> 
    ggplot(aes(x=k)) + 
    geom_line(aes(y=training_rmse), color="red4") + 
    geom_line(aes(y=test_rmse), color="blue4") + 
    geom_line(aes(y=validation_rmse), color="green4") + 
    ggtitle("Comparison of Errors: Training (Red), Test (Blue), Validation (Green)")
```

So the validation error is consistently a bit over pessimistic, but generally a better guess as to the shape of the true test error. 

Why is the green error consistently a bit too high here? We're only
using 32 instead of 40 data points, so things are a bit worse, but
with a larger data set, the effects might not be so profound. 

But going to 32 data points was _expensive_ - not computationally, but
statistically. We definitely wound up comparing worse models. Could we
do better? 

What if we could hold out only 1 data point? The model would be better, but
the error measure would be worse... But we could repeat this process for
each point and aggregate (average) our results across all 40 fits. 

```{r}
make_leave_one_out_predictions <- function(k){
  N <- NROW(OBSERVED)
  
  map(seq(N), \(N){
      OBSERVED_LESS_ONE <- OBSERVED[-N, ]
      HELD_OUT_POINT <- OBSERVED[N, , drop=FALSE]
      
      knn_k <- nearest_neighbor(neighbors = k, 
                          # No weights - treat all neighbors equally
                          weight_func="rectangular") |>
        set_engine("kknn") |>
        set_mode("regression") |>
        fit(Y ~ X, data = OBSERVED_LESS_ONE)
  
      augment(knn_k, new_data=HELD_OUT_POINT) |> 
      rename(Y_hat = .pred) |>
      mutate(k=k)
  }) |> bind_rows() |>
      summarize(loo_rmse = sqrt(mean((Y - Y_hat)^2)), 
                k=first(k))
}

map(1:25, make_leave_one_out_predictions) |>
    bind_rows() |> 
    ggplot(aes(x=k, y=loo_rmse)) + 
    geom_point() + 
    xlab(expression(K)) + 
    ylab("Leave One Out RMSE") + 
    theme_bw() + 
    ggtitle("Selecting the K in KNN via LOO")
```
