---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}


Outline: 

--- 

So far, we have talked mainly talked about models and different
aspects of understanding predictive performance (in/out-of-sample,
test/train, decomposing error into sources). Throughout all of this, we've
repeatedly made reference to the empirical risk minimizer: 

$$\hat{f} = \argmin_{f \in \Fcal} \frac{1}{n}\sum_{i=1}^n L(f(\bX_i), Y_i) \tag{ERM}$$

Today, our focus is on actually _solving_ the ERM problem. This can get quite
technical, so we're just going to work on building the intuition for how one
could solve this type of problem. 

The first-and most important-thing to note is that the ERM problem essentially
never has a simple closed form solution: OLS (and some close cousins)
are a notable exception. Even something as simple and familiar as _logistic 
regression_ requires us to adopt an alternative approach. 

To keep our notation simple, let's rewrite the ERM problem as a special case of
a more general problem: 

$$x_* = \argmin_{x \in \Ccal} f(x)$$

That is, we want to find the value(s) of $x$ that make $f(\cdot)$ as small
as possible. Obviously this formulation captures ERM, but it also includes 
problems in applied math as vast as transportation planning (minimize 
fuel usage), economic decision making (minimize costs associated with a 
complex decision), classical statistics (minimize missfit between a model and
some observed data) and many others. The function $f$ is called the
_objective_. Note that we're not loosing anything by focusing on _minimizing_
the objective. It's easy to see that

$$\argmin_{x \in \Ccal} f(x) \Leftrightarrow \argmax_{x \in \Ccal} -f(x)$$

The requirement $x \in \Ccal$ is intentionally quite vague: we're simply
requiring $x$ to live in _some set_, traditionally known as a _constraint_ set.
In the simplest problems, $\Ccal$ is all vectors (all possible combinations of
real numbers) and we say the problem is _unconstrained_. In other problems, 
we might use $\Ccal$ to impose constraints that make a solution workable - we 
minimize fuel cost _subject to actually getting to our destination_ and we
minimize costs associated with a complex decision _subject to contracts we 
have negotiated in the past_. 

In ML, we rarely have these types of hard domain-specific constraints, but we 
will want to use the constraint set to _nudge_ our answers in a particular 
direction. (We will return to this when we talk about regularization in Unit II.)

For now, let's assume we have no constraints, so we are just solving 

$$\argmin_{x \in \R^p} f(x)$$

## Why Not Grid Search?

## Unconstrained Optimization via Gradient Descent

All of modern ML optimization relies on one remarkable fact from elementary
calculus: the derivative points the way. 

Let's go back to basics: essentially all functions in ML take in several 
inputs (parameters) and produce a single output. Formally, 

$$ f: \R^p \to \R$$

The relevant notion of derivative for this type of 'scalar-valued' function $f$
is the _gradient_: the gradient, which we denote $\nabla f$ is a _vector_ of
derivatives. If we think about a two-input function $g(x, y)$, its gradient
is the vector given by 

$$\left(\frac{\partial g}{\partial x}, \frac{\partial g}{\partial y}\right)$$

That is, we take the derivative with respect to _each_ input one at a time and 
join them into a vector. Why is this useful? 

Remember how a vector works in simple single-variable calculus: it 
is _positive_ if the function is increasing at a point and negative if the 
function is decreasing. Another way to say this is: "the derivative is positive
if moving in the positive direction (rightwards) makes $f$ go up and the 
derivative is negative if moving in the negative direction (leftwards) makes $f$
go up."

Visually, 

```{r}
#| message: false
#| echo: false
#| fig-asp: 1

x <- seq(-3, 3, length.out=201)
f <- x^2 + 3 * x + 2

plot(x, f, type="l", xlab="x", ylab="f(x)")

# Draw tangent line at x=1
abline(b=5, a=1, col="red4")
text(2, 6, "Direction of Increase ->")
```

and 

```{r}
#| message: false
#| echo: false
#| fig-asp: 1

x <- seq(-3, 3, length.out=201)
f <- log(3-x)

plot(x, f, type="l", xlab="x", ylab="f(x)")

# Draw tangent line at x=1
abline(a=log(2) + 0.5, b=-0.5, col="red4")
text(2, 1, "<- Direction of Increase")
```

This suggests an approach we can use to find the minimum of any function: 

- Given a starting point $x_0$:
- Repeat until the minimum is found: 
    - Compute the derivative $f'(x_k)$
    - If positive, shift a bit to the left $x_{k+1} = x_k - \epsilon$
      If negative, shift a bit to the right $x_{k+1} = x_k + \epsilon$
      
We can repeat this simple loop and eventually run into **a** minimum of the 
function. Visually, suppose we want to find the minimum of this function: 

```{r}
f <- function(x) (x + 2) * (x + 1) * (x - 1) * (x - 3) + 0.5
curve(f)
```

Let's start at $x = 0$ and see what happens: 

```{r}
# TODO
```

This algorithm can be generalized to give us the general strategy of 
_gradient-descent_: we repeat the following loop until we reach a minimum: 

$$x^{(k+1)} \leftarrow x^{(k)} - c_k \nabla f(x^{(k)})$$

Notably, when the derivative is zero $\nabla f = \bzero$ we stop. (Check why.)

*Aside*: If we were _maximizing_, we would have $x^{(k)} + c_k \nabla f$. Why?

This strategy has several very nice properties: 

- It gives us a natural stopping rule, at least in theory, of $\nabla f = \bzero$. 
  In practice, getting an _exact_ zero is very hard, so we typically stop when
  $\nabla f \approx \bzero$ or when $f(x^{(k+1)}) \approx f(x^{(k)})$
- As long as the step size $c_k$ is chosen small enough, we always are going
  to go downhill ($f$ decreases). 
- We don't need any _global_ information. We only need the gradient at one point
  at a time. We sometimes say we only need a first-order _oracle_. We don't
  really need to know the _function_ $\nabla f$: we only need to be able to
  evaluate it. For very complex functions, this turns out to be a real practical
  advantage as we will discuss below. 

This last point is a bit of a double edged sword: gradient descent will help us
find a point where $\nabla f \approx \bzero$, but this only guarantees us a
a _local_ minimum (or possibly even just a critical point). There may be a 
better solution farther away that we don't see. *E.g.*, 

```{r}
#| echo: false
f <- function(x) (x - 3) * (x-1) * (x + 1) * (x+4)
curve(f, from=-5, to=5, main="A Function With Two Minima")

points(-2.94, f(-2.94), col="red4", pch=16, cex=2)
points(2.225, f(2.225), col="blue4", pch=16, cex=2)
```

Here, we see that we have _the_ minimizer, also called the _global
minimizer_ at the red dot as well as a _local_ minimizer at the
blue dot. If we start GD around the blue dot, we will 'fall down'
into that valley and converge to the blue dot, but we will never get
to the true global minimizer. GD cannot easily escape this sort of
trap because it only looks at the derivative around the blue dot
and there is no local direction to go downwards. 

### Learning Rates

In our basic GD update

\[x^{(k+1)} \leftarrow x^{(k)} - c_k \nabla f(x^{(k)})\]

what role is played by the value $c_k$? We haven't really talked about
it so far. 

$c_k$ is known as the _step-size_ (in general optimization theory) or
the _learning rate_ (in ML). It controls _how fast_ we go in GD. If
we pick a larger $c_k$, we take proportionally larger steps. 

Big steps seem great - they will help us get to our destination 
(minimizer) more quickly. And sometimes they will. But they pose
a bit of a risk. If our step size is _too_ large, we can 'step over'
the minimum. 

Take a very simple example: minimizing $f(x) = x^2$, where the answer
is clearly $x = 0$ and we have $f'(x) = 2x$. If we start at 
$x^{(0)} = -2$ and take a simple step size of $c_k = 1$, we will just
have: 

$$\begin{align*}
x^{(0)} &= -2
x^{(1)} &= x^{(0)} - c_k f'(x^{(0)}) \\
        &= -2 - 1 * (2 * (-2)) \\
        &= 2 \\
x^{(2)} &= x^{(1)} - c_k f'(x^{(1)}) \\
        &= 2 - 1 * (2 * 2) \\
        &= -2
x^{(3)} &= x^{(2)} - c_k f'(x^{(2)}) \\
        &= -2 - 1 * (2 * (-2)) \\
        &= 2
\dots &= \dots
\end{align*}$$

and we will 'ping-pong' back and forth forever. This example is
a bit contrived and a perfect cycle is pretty rare, but the general
phenomenon of bouncing around 'above' the valley certainly happens. 

So what can we do? The simplest choice might be to set $c_k$ to some
small value, but this comes with its own costs. We can take smaller
steps, but we will then need many more to get to our minimizer. 

One simple strategy is to have $c_k$ not be constant, but to instead
have it _decay_ in $k$: that is, to have the step sizes get smaller
as we go on. A simple choice might be $c_k = 1/k$. We start with
a bit step $c_1 = 1$, but our next step is half-sized $c_2 = 1/2$
and so forth. Our hope here is to take an initial few large steps
to get us close to the minimizer and then to 'fine-tune' our way to
the actual minimizer with smaller steps. 

This does work, but it can be a bit finicky. If we have $c_k$ decay
too quickly, we get ourselves back to the small-steps / many-steps
problem (even more so if $c_k$ gets _very_ small); but if $c_k$
decays too slowly, we can waste a lot of time ping-ponging. 

This motivates a _huge_ variety of different procedures for
picking $c_k$ and much of applied ML research revolves around coming
up with new variants. This is one of those parts of ML where a 0.01%
improvement can be worth _billions_ of dollars because it makes
_every_ training step more efficient. 

Generally, rules for picking $c_k$ fall into two camps: 

- _Momentum_: if we have taken several steps in the same direction, 
  we should assume we want to keep going in that direction and keep
  our step size large. 
- _Adaptivity_: if the function is changing slowly, we are in a 
  'flat region' and probably want to take big steps; if it is changing
  quickly, we want to slow down so we don't over step. 
  This is particularly useful so that we can _adapt_ to the type of
  function as opposed to having to know up front whether it is a
  'big step' or 'small step' function. 
  
I like to think of these as how one might descend a mountain. In the
flat parts, we can take big steps without any fear; in the steep
parts, we take smaller steps so we don't fall flat on our face. 

Modern ML optimization methods combine these two ideas - and several
others - to try to make sure that we are always taking large, but
not too large, steps in a useful direction. It's a bit dated now, but
for many years the `Adam` optimizer dominated this space, as it was
one of the first to combine `Ada`ptivity and `M`omentum in an easy
to use package. 

### AutoDiff

So we know we can use a learning rate given to use using an Adam-type
rule in our GD update: 


\[x^{(k+1)} \leftarrow x^{(k)} - c_k \nabla f(x^{(k)})\]

We're still left with one _very painful_ chore: computing the
gradient $\nabla f$ at each step. Recall that our goal was to not
have to evaluate $f$ too many times and now we're asking for a 
derivative. 

We have two obvious ways of doing this, both of which stink: 

- Actually do the symbolic calculation of computing $\nabla f$,   
  *i.e.*, do some calculus by hand. This works for toy problems
  (as shown above), but it's a huge pain for realistic models. 
  In many circumstances, we don't even have $f$ as a single formula
  so differentiating it manually is going to be ludicrously terrible.
- Using a difference quotient. You might remember from calculus that
  
  $$f'(x) = \lim_{\Delta x \to 0} \frac{f(x + \Delta x) - f(x)}{\Delta x}$$
  
  So, if we can't find $f'$ directly, if we compute $f$ twice at
  slightly different values of $x$, we can get an approximate
  (numerical) derivative. This is a reasonable strategy in
  one-dimension, but it doesn't scale up to big complex models. 
  
  Recall that $\nabla f$ is the derivative in all $p$-dimensions, so
  we would need to evaluate $f$ $p+1$ times to get a single
  derivative. At this point, we're starting to get back into the
  grid search scale of evaluating $f$. 
  
A remarkable idea lets us get around this: at the end of the day, 
computers can only do a few very simple operations: 

- They can add and subtract
- They can multiply and divide
- They can call other functions

Everything else is built out of these: if we compute a least squares
loss, there's a bunch of matrix multiplications, squaring, and
summing, but these are just the same basic parts. Moving to a more
complex loss, say one with a logarithm, might seem tricky, but
under the hood, computers just use a power series to compute a
logarithm anyways. There is no logarithm-unit in hardware. 

But these are all things where we understand the derivative well. From
Calculus I:

$$\begin{align*}
(f + g)' = f' + g' \\
(f*g)' &= f'*g + f * g' \\
(f \circ g)' &= f'(g) * g'
\end{align*}$$

If any compute program is built out of these pieces - admittedly at
_enormous_ scale - can't we just use them to compute the derivative?
In fact, we can, with lots of book-keeping. But book-keeping is
the sort of things computers are great at!

This gives rise to _automatic-differentiation_ or _auto-diff_. 
(Sometimes also called _back-propagation_ or _back-prop_ in reference
to the core book-keeping algorithm.) Auto-diff is the backbone of 
most modern ML toolkits: we simply code up $f$ and through the
magic of auto-diff, we get the ability to execute $\nabla f$ 
'for free'.[^free]

To wit: taking 

$$f(x) =  (x - 3) * (x-1) * (x + 1) * (x+4)$$

```{python}
import jax

def f(x):
    return (x - 3) * (x - 1) * (x + 1) * (x + 4)

x = 2.0

# --------------------------------------------------
# 1. Using jax.grad:
# Requires calling f(x) and df(x) separately
# --------------------------------------------------
df = jax.grad(f)

val = f(x)
grad = df(x)

print("jax.grad:")
print(f"  f({x})  = {val}")
print(f"  f'({x}) = {grad}\n")

# --------------------------------------------------
# 2. Using jax.value_and_grad:
# Computes both in a single call
# --------------------------------------------------
f_and_df = jax.value_and_grad(f)

val, grad = f_and_df(x)

print("jax.value_and_grad:")
print(f"  f({x})  = {val}")
print(f"  f'({x}) = {grad}")
```

Here, we see that we can either just use `jax.grad` to automatically
create a new derivative function or we can even use
`jax.value_and_grad` to get them both at once. In both cases, $f$ 
is the input and we get back a _new_ function which is then later
evaluated at a point. 

We can use this to create a simple gradient descent loop: 

```{python}
import jax
import jax.numpy as jnp

def f(x):
    return (x - 3) * (x - 1) * (x + 1) * (x + 4)

class GradientDescentUpdate:
    def __init__(self, objective, init_x, learning_rate):
        self.learning_rate = learning_rate
        self.x = jnp.array(init_x, dtype=jnp.float32)
        self.grad_fn = jax.grad(objective)

    def step(self):
        grad = self.grad_fn(self.x)
        self.x = self.x - self.learning_rate * grad
        return float(self.x), float(grad)

initial_x = -4.0

opt = GradientDescentUpdate(
        objective=f, 
        init_x=initial_x,
        learning_rate=0.0001 # Need this to be _very_ small since not changing!
)

# Outer loop managed explicitly by caller
num_steps = 50
tol = 1e-5

old_x_val = initial_x

for step in range(num_steps):
    new_x_val, grad = opt.step()
    print("Updated x to")
    print(new_x_val)
    if abs(grad) < tol or abs(new_x_val - old_x_val) < tol:
        print(f"Converged at step {step}")
        break
    old_x_val = new_x_val

print(f"Final x: {float(opt.x):.4f}")
print(f"Objective value: {f(opt.x):.4f}")
```

Ok - not exactly correct and maybe a bit clunky - but this basically worked!
There is a minimum at $x=-4$ and we got pretty darn close. 

[^free]: Not entirely _for free_ because we do have the book-keeping
steps, but when done smartly, this can be something like 8-10x as slow
as just computing $f$. Certainly not nothing, but much better than
computing $f$ $p$ different times in a $p = 10^8$ parameter model. 

### Convexity

So returning to our example from above, how can we _guarantee_ that we get
the _global_ optimum (red dot) instead of just a local optimum (blue dot)? 
Looking at the basic GD structure, it's clear that if we start sufficiently 
near the global optimum, we'll find it, but it's not at all obvious how
near is near enough or how we can ever be sure. 

This is a **really** hard problem in general, but there are a useful class
of functions where we can be guaranteed to get the _global_ minimum. Before
saying _how_, it's important to pause and ask _why_ it may (or may not) be
important to get the global minimum. 

1) To an academic, much of our theory applies to the global minimum. Whether
   this is important or not depends on how much you value theory. 
2) There is a reproducibility argument: if we _always_ get the global minimum,
   then choices like learning rate or starting point don't matter, which
   is great. (Good life advice: it's best if hard choices don't actually change
   the outcome.)
3) There is also a 'statistical' stability argument. If we think of minimizing
   a loss function based on _noisy_ data, it's probably a good thing if the
   minimizer doesn't move very much. If there are multiple minimizers 
   (as in our blue/red example above), a small change in the function can
   totally change the location of the global minimizer; this might render
   our models unstable or unpredictable. 
   
So how can we guarantee this sort of _global_ optimality. Well, like most
things in math, we cheat a bit: specifically, we just assume that all
_local_ minima are _global_ minima. Of course, we're not so explicit about
that. We instead say that we are guaranteed to get the global minimizer if
$f$ is _convex_, *i.e.*, bowl shaped. 

For a convex function like: 

```{r}
#| echo: false
f <- function(x) (x - 2) * (x + 1)
curve(f, from=-3, to=3)
```

There is only one minimizer and it is both local and global. So finding
a local minimizer means we are guaranteed to get a global minimizer. 

Not all convex functions have a single minimizer: 

```{r}
#| echo: false
f <- function(x) pmax(x^2 - 3, 0)
curve(f, from=-3, to=3)
```

Here we have many minimizers (that whole flat region), but they are _all_
giving the global optimum ($f_* = 0$). 

Formally, a function $f$ is convex if: 

$$f(\lambda x + (1 - \lambda)y) \leq \lambda \lambda f(x) + (1 - \lambda) f(y) \text{ for any } \lambda \in [0, 1]$$

That's a bit formal, but what does it mean. It means that if we pick
any two points $x, y$ and draw the line connecting them ($lambda f(x) + (1 - \lambda) f(y)$), $f$ lies below it

In our example above: 


```{r}
#| echo: false

f <- function(x) (x - 2) * (x + 1)
curve(f, from=-3, to=3)

points(1.5, f(1.5), pch=15, cex=2, col="red4")
points(-1.5, f(-1.5), pch=15, cex=2, col="red4")

segments(-1.5, f(-1.5), 1.5, f(1.5), lty=2, col="red4")
```

*TODO*: Why does this guarantee optimization will work nicely? Given two guesses, we can get a better guess by looking between them. 



## Stochastic Gradient Descent

## Zeroth Order Optimization
