---
notes_num: "`r n <- knitr::current_input() |> stringr::str_extract('notes([:digit:]+).(qmd|rmarkdown)', group=1) |> as.integer(); sprintf('%02g', n)`"
date: "`r Sys.Date()`"
notes_title: "`r readr::read_csv('key_dates.csv') |> dplyr::filter(Item == 'Class Session', Number == n) |> dplyr::pull(Details)`"
---

{{< include _latex_defines.qmd >}}

Outline: 

--- 

## Challenges of High-Dimensional Regression

Last time, we discussed several challenges associated with doing
statistical inference with _many_ features; that is, data that is
_high-dimensional_. Principally, these came from two sources: 

- Multiple testing problems: if we take enough swings, any _rare_
  event will eventually occur. In the context of statistical testing, 
  this can blow up our normal error rate guarantees. Naively, we 
  can just raise our bar to be ever stricter, but without sufficient
  data, this just winds up putting everything 'out of reach'
- Large variance problems: when estimating _many_ values, the variance
  of our estimates can become uncomfortably large. To avoid this 
  variance, we are willing to take on a bit of _bias_ and, we balance
  things out correctly, this can net out well for us. 
  
  This bias can take two forms: 
  
  - Intuitively, if we have lots of related little problems, we can
    partially _pool_ information, *e.g.*, shrink everything a bit
    towards the grand / population mean. Here, we massively reduce
    variance by blending together a large number of samples into the
    'shrinkage target' and, if those samples are not too unrelated, 
    we don't incur too much bias along the way. 
  - Less intuitively, we don't actually need our different problems
    to be related at all; even in the case of estimating many
    independent normal means, a _little_ bit of shrinkage always
    helps. 
    
That latter phenomenon - most notably explored by James and Stein - is
and should be quite counterintuitive. But ultimately, it comes from a
close examination of bias and variance. 

If we are estimating $p \gg 1$ different means from a set of sample
means, each with variance $\sigma^2$, the total MSE of our problem
will be $p \sigma^2$. (Sample means are unbiased, so we only have
variance here and variance adds for independent variables.) If instead
of using the sample means $\hat{\mu_i}$, we instead look at a slightly
'shrunk' version $(1 - \rho)\hat{\mu_i}$, we now have a small bias of 

$$\text{Bias}_i = \mu_i^* - \E[(1-\rho)\hat{\mu_i}] = \mu_i^* - (1-\rho)\mu_i^* = -\rho\mu_i^*$$

and a reduced variance of 

$$\text{Variance}_i = \V[(1-\rho)\hat{\mu_i}] = (1-\rho)^2\V[\hat{\mu_i}] = (1-\rho)^2 \sigma^2$$

Putting these together, the MSE for each individual sample is 

$$\text{MSE}_i = \text{Bias}_i^2 + \text{Variance}_i = \rho^2(\mu_i^*)^2 + (1-\rho)^2\sigma^2$$

A bit of calculus shows that the _best_ $\rho$ we could pick is 

$$\rho_i^* = \frac{\sigma^2}{(\mu_i^*)^2 + \sigma^2}$$

which gives an MSE of 

$$\text{MSE}(\rho_i^*) = TODO$$

Of course, getting this right required us to know $\mu_i^*, \sigma^2$ 
which we don't normally have. (If we knew $\mu_i^*$, why are we doing
any of this?)

James and Stein's remarkable finding was to show that a slight variant
of this process works _for all_ $\mu_i^*, \sigma^2$. 

\[\hat{\mu}_i = TODO \]

What does any of this have to do with regression? Well, let's go back
to that MSE formula above; instead of picking just the best $\rho$,
let's look at the _derivative_ at 0: 

\[\frac{\text{d}\,\text{MSE}}{\text{d}\,\rho} = \rho(\mu^2 + \sigma^2) - \sigma^2 \implies \left.\frac{\text{d}\,\text{MSE}}{\text{d}\,\rho}\right|_{\rho = 0} = -\sigma^2 < 0\]

So a _little_ bit of shrinkage towards zero always helps. The ideal
amount depends on the specifics of $\mu, \sigma$, but if there is any
noise, a little bit of shrinkage towards zero should always help us.

Just in case you don't believe me: 

```{r}
library(tidyverse)
library(ggforce)

mu <- 3
sigma <- 1
estimate_mse <- function(rho){
    X <- rnorm(5000, mean=mu, sd=sigma)
    shrunk_estimate = (1 - rho) * X
    data.frame(rho, 
               mse_shrinkage = mean((mu - shrunk_estimate)^2))
}

simulation_mse <- 
    map(seq(-2, 2, length.out=5001), estimate_mse) |>
    list_rbind()


ggplot(simulation_mse, 
       aes(x=rho, y=mse_shrinkage)) + 
    geom_point() + 
    theme_bw() + 
    xlab(expression(rho)) + 
    geom_hline(yintercept=estimate_mse(0)$mse_shrinkage, 
               color="red4") + 
    ylab("MSE") + 
    ggtitle("Effect of Shrinkage on Estimating Sample Means") + 
    facet_zoom(xlim=c(-0.1, 0.1), 
               ylim=c(0, 1))
```

Admittedly, this isn't the easiest to see and we had to zoom, but
it's there! At zero, there is a small downward slope, so a tiny bit 
of shrinkage helps, even in this super simple problem. 

In this way, I like to think of shrinkage as like _salt_ in cooking.
A little bit is essentially always helpful, but too much can certainly
be a problem.

This general phenomenon is also referred to as **regularization**, as it relates
to an older mathematical idea of 'smoothing out' the rough edges of models to
make them more tractable. We won't use that perspective in this course, but 
the term **regularization** is definitely worth knowing. 

### Ridge Regression - Core Idea

Ridge regression extends this idea to the high-dimensional regression
setting. Recall that by high-dimensional, we many data with _many_
features, possibly even more features than there are samples in the
data set ('wide' data). In these scenarios, we have _lots_ of variance
--because we have little data to 'tame' it--so we want to apply
some shrinkage to our data. 

While the 'make everything a bit smaller' strategy does work, it's a
bit more conventional to represent this in a different way: we choose
to keep the regression coefficients from being _too big_. Specifically,
we're going to put a _max size_ on the _total_ of our coefficients. 

By imposing a 'budget' on our problem, we're going to force most of
the small effects to be even closer to zero and prevent the big effects
from being _too big_. (Or at least, that's the hope.) We can think of
this as a sort of 'adaptive' shrinkage. While we started by shrinking
multiple things totally independently, regression 'blends' all of our
features together, so we want a more gloabl sort of shrinkage. 

## Constraints

To do this, we need to introduce the idea of _constrained_ ERM. 
Recall that our general ERM for linear regression under MSE loss was

$$\hat{\bbeta} = \argmin_{\bbeta \in \R^p} \frac{1}{n}\sum_{i=1}^n (y_i - \bx_i^{\top}\bbeta)^2 \tag{OLS}$$

Here, we only require $\bbeta$ to be a set of $p$ real-valued
coefficients (for our $p$ features). This problem is functionally
unconstrained. But we might consider other sorts of constraints:
*e.g.*, we might know that our features could only have a non-negative
(zero or positive) coefficient based on some prior knowledge about
the problem. In this case, we could modify the above problem as

$$\hat{\bbeta} = \argmin_{\substack{\bbeta \in \R^p \\ \beta_i \geq 0 \text{ for all } i}} \frac{1}{n}\sum_{i=1}^n (y_i - \bx_i^{\top}\bbeta)^2 \tag{NN-OLS}$$

This gives us _non-negative_ OLS. Instead of looking at all coefficients, we 
just find the best _non-negative_ $\hat{\bbeta}$. That second part: non-negativity
is a constraint. You can also write this as $\bbeta \in \R^p_{\geq 0}$ where 
$\R^p$ is the 'positive orthant' (the 'positive corner' of the real plane). 
Because $\R^p_{\geq 0} \subsetneq \R^p$, this is a _constrained_ problem. 

Let's think about what this does: 

- We're looking over a smaller search space, so we have less variance
- **If** our true $\bbeta^*$ has some negative values, we will be
  systematically wrong, introducing bias. 
  
So this is one of those 'bias-variance' gambits. If our restriction
to non-negative values is true, this is a pure-win on variance and no
pain on bias; if our true values are very large and negative, this can
hurt us quite a lot. And if our true values are a mix of near zero, 
positive and negative, we may or may not benefit. 

Alternatively, we can note that this is just like any sort of assumption in 
our modeling: if it's true, it helps; and if it's false, it can hurt. 

Finally, we note that we might have other reasons beyond pure MSE optimization
to consider this type of thing: NN-OLS may be far simpler to interpret 
and, it turns out, will wind up 'zeroing' (throwing out) lots of features. But
we're getting a few days ahead of ourselves. 

### Measuring the Size of Coefficients

Ok - non-negativity is straightforward enough. But how do we encode something
like 'not too big'? We need a way of describing how big a set of regression
coefficients is. 

Which of these is a 'bigger' set of coefficients: 

$$(1, -1, 1) \text{ vs. } (2, 0, 0)$$

There are clearly several ways of measuring bigness that might make sense: 

- Number of non-zero elements (number of features used). Here, the first
  vector is clearly 'bigger' (3 vs 1)
- Maximum (absolute) coefficient. Here, the second vector is clearly bigger
  (2 vs 1)
- Sum of absolute coefficients. Here, the first vector edges out the
  second $|1| + |-1| + |1| = 3 > 2 = |2| + |0| + |0|$.
- Sum of squared coefficients. Here, the second edges out the first: 
  $2^2 + 0^2 + 0^2 = 4 > 3 = 1^2 + (-1)^2 + 1^2$
  
All of these are correct for different types of problems, but we need a
mathematical way to distinguish them. The mathematical concept of a **norm**
measures 'how big' something is and, when applied to vectors of coefficients,
we can measure these using the $\ell_p$ family of norms: 

$$\|\bx\|_p = \sqrt[p]{\sum_{i=1}^n |x_i|^p}$$

So, with our examples above, for the $\ell_1$ norm, we had: 

$$\|(1, -1, 1)\|_1 = \sqrt[1]{|1|^1+ |-1|^1 + |1|^1} = 3$$

and 

$$\|(2, 0, 0)\|_2 = \sqrt[2]{|2|^2+ |0|^2 + |0|^2} = 2$$

Here, the $\ell_2$ norm measures distance 'as the crow flies' or in a straight
line. (You might notice that this looks sort of like the Pythagorean theorem; 
that's no coincidence.) Conversely, the $\ell_1$ norm captures how many steps
you might need to take on a rectangular grid. For this reason, the $\ell_2$ norm
is sometimes called the _Euclidean_ norm while the $\ell_1$ is called the
_taxicab_ or _Manhattan_ norm because it captures how many blocks you would need
to walk.[^blocks]

Visually, these constraints give us different 'balls' that we are requiring
our coefficients to fall in: 

```{r}
#| echo: false
#| fig.asp: 0.3
library(patchwork)
plot_ball <- function(p){
    title <- bquote(l[.(p)]~"Unit Ball")
    expand_grid(x=seq(-2, 2, length.out=201), 
                y=seq(-2, 2, length.out=201)) |>
        mutate(interior = abs(x)^p + abs(y)^p <= 1) |>
        filter(interior) |>
        ggplot(aes(x=x, y=y)) + 
        geom_point() + 
        coord_fixed() + 
        ggtitle(title) + 
        theme_bw() + 
        xlab("") + ylab("")
}
plot_ball(1) + plot_ball(2)
```

Here, the $\ell_1$ "ball" is pointy and isn't a ball in any normal sense, but
that's the terminology of math. The $\ell_2$ ball should feel familiar.[^ped]

[^ped]: Formally, the $\ell_2$ 'circle' is the boundary while the entire
enterior is the 'ball'. We basically always use 'ball' constraints instead
of circle constraints. If the best value is even smaller than the edge of the
ball, we don't want to 'drag' things outwards. 

![From Wikipedia on [Taxicab Geometry](https://en.wikipedia.org/wiki/Taxicab_geometry)](https://en.wikipedia.org/wiki/Taxicab_geometry#/media/File:Manhattan_distance.svg)

[^blocks]: Yes, East-West and North-South blocks are not the same size in
Manhattan, so the 'real' Manhattan norm should treat the two dimensions
differently, but Manhattan isn't $p$-dimensional so we accept a bit of 
over-simplification. 

To set up our first form of *regularized* regression, we're going to use
the $\ell_2$ norm: 

$$\hat{\bbeta} = \argmin_{\bbeta: \|\bbeta\|_2^2 \leq \tau^2} \frac{1}{n}\sum_{i=1}^n L(\bbeta^{\top}\bx_i, y_i) \tag{Ridge Regression}$$

Here, we're requiring the sum of squared coefficients from being more than
some value $\tau$. $\tau$ here is a _hyperparameter_ that we have to choose. 
Bigger $\tau$ gives us a larger set of possible $\bbeta$, hence more variance
and less bias, while smaller $\tau$ gives us more bias, but less variance.[^sd]

[^sd]: Remember that the standard deviation of _anything_ in statistics that
takes values falling in the range $[a, b]$ is never more than $|b-a|/2$, so
by keeping our values in a $[-\tau, \tau]$ range, the standard deviation is
basically capped at $\tau$. 

How can we choose $\tau$? This, like the $K$ in $K$-NN, is one of those 
situations where pure ERM will fail us. Since the ball associated with a
larger $\tau$ always includes a smaller $\tau$, pure _in-sample_ minimization
will always push us towards a bigger $\tau$. We need to use some sort of
holdout / cross-validation strategy here and will return to that in the future. 

### Constraints and Penalties

Above, we write ridge regression as: 

$$\hat{\bbeta} = \argmin_{\bbeta: \|\bbeta\|_2 \leq \tau} \frac{1}{n}\sum_{i=1}^n L(\bbeta^{\top}\bx_i, y_i) = \argmin_{\bbeta: \|\bbeta\|_2 \leq \tau} \frac{1}{2}\|\by - \bX\bbeta\|_2^2$$

which is essentially the classical OLS problem (MSE ERM with a linear
predictor) with the additional constraint: $\|\bbeta\|_2 \leq \tau$ or,
slightly more conveniently, $\|\bbeta\|_2^2 \leq \tau^2$.[^sq] Here,
we're putting the 'ridge-ness' in the _constraint_ by requiring $\bbeta$ to not
be 'too big.'

This is conceptually elegant, but it does pose a few practical challenges: 

- How can we pick $\tau$? 
- How can we solve the above problem? In particular, what happens if our
  gradient points us 'beyond the bounds' of our constraint? 
  
These are definitely solvable (and indeed solved!) problems, but it certainly
would be nice if we could get rid of the constraint. And as it turns out, 
we can. A bit of mathematical trickery lets us rewrite the above problem as


$$\hat{\bbeta} =\argmin_{\bbeta: \|\bbeta\|_2 \leq \tau} \frac{1}{2}\|\by - \bX\bbeta\|_2^2 = \argmin_{\bbeta \in \R^p} \frac{1}{2}\|\by - \bX\bbeta\|_2^2 + \lambda \|\bbeta\|_2^2$$

Here, we have moved our 'ridgefulness' out of the constraint and into a second
part of our objective function. This is known as a **penalty** term as it 
keeps $\bbeta$ from being 'too big'. Notably, unlike the constrained form, 
we could theoretically get a _really big_ $\bbeta$, but it would have to
'pay for itself' in terms of better MSE. Essentially, we are putting a 'price'
on bigger coefficients. We're willing to use them, but only if they 'prove their
worth' in terms of MSE.[^stepwise]

Constraints are a bit of a pain computationally (we have to modify normal 
gradient descent so it doesn't take out out of bounds), so this _penalized_ form
is a bit more common in practice, but there is no conceptual difference. The
penalty form lets us see how this works quite nicely. 

Suppose we were using a first-order (gradient descent method). Normally, we would
just take a step in the direction of the gradient of the loss. What happens here?
Since the gradient of $\|\bbeta\|_2^2$ always points outwards, a _descent_ step
will always steer us a bit back towards the origin. 

- Small effects => pirhanas, why not choose?
- Small effects => stability
- Weight decay

[^stepwise]: If this sounds like a 'soft' version of using $F$-tests, ANOVAs, 
or stepwise methods to decide what features to include in model, you aren't wrong.


[^sq]: Any norm is non-negative, so both terms of this inequality are
non-negative and squaring doesn't change the order.

## Ridge Regression


**These notes are from a previous class I've taught and are a bit beyond the 
level of this course.**

In the [previous set of notes](./notes06.qmd), we saw that
$(\overline{X}_n)_{[-\beta, \beta]}$ could outperform the naive estimator
$\overline{X}_n$ if the true parameter is 'not too big'. 

Can we extend this idea to regression? 

Recall that we formulated OLS as: 

$$\hat{\bbeta}_{\text{OLS}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\|$$

We can apply a 'clamp' as: 

$$\hat{\bbeta}_{\tau-\text{Clamped}} = \argmin_{\bbeta \in \R: \|\bbeta\| < \tau} \frac{1}{2}\|\by - \bX\bbeta\| = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\| \leq {\tau}$$

Here, we have some choice in measuring the 'size' of $\hat{\bbeta}$: in fact, we
can *theoretically* use any of our $\ell_p$-norms. As with the squared loss, 
it turns out to be mathematically easiest to start with the $\ell_2$ (Euclidean)
norm. This gives us the so-called *ridge regression* problem: 

$$\hat{\bbeta}_{\text{Ridge}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2 \leq {\tau} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2^2 \leq {\tau}^2$$

Before we discuss solving this, let's confirm it is convex. Recall that for a
problem to be convex, it needs to have a convex *objective* and a convex *feasible
set*. Clearly the objective here is convex - we still have the MSE loss from OLS
unchanged. So now we need to convince ourselves that $\|\bbeta\|_2^2 \leq \tau^2$
defines a convex set.  We can look at this in 2D first: 

$$\|\bbeta\|_2^2 \leq \tau^2 \implies \beta_1^2 + \beta_2^2 \leq \tau$$

But this is just the equation defining a circle and its interior (in math speak
a 2D 'ball') so it has to be convex!

In fact, constraints of the form 

$$\|\bbeta\|_p \leq \tau \Leftrightarrow \|\bbeta\|_p^2 \leq \tau^p$$

are convex for all $\ell_p$ norms ($p \geq 1$). We will use this fact many times
in this course. The sets that this constraint defines are called $\ell_p$ 'balls'
by analogy with the $\ell_2$ figure, even if they aren't actually round.

```{r}
#| echo: false
#| fig.asp: 0.3
library(patchwork)
plot_ball <- function(p){
    title <- bquote(l[.(p)]~"Unit Ball")
    expand_grid(x=seq(-2, 2, length.out=201), 
                y=seq(-2, 2, length.out=201)) |>
        mutate(interior = abs(x)^p + abs(y)^p <= 1) |>
        filter(interior) |>
        ggplot(aes(x=x, y=y)) + 
        geom_point() + 
        coord_fixed() + 
        ggtitle(title) + 
        theme_bw() + 
        xlab("") + ylab("")
}
plot_ball(1) + plot_ball(2) + plot_ball(Inf)
```

So our ridge problem 

$$\hat{\bbeta}_{\text{Ridge}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2^2 \leq {\tau}^2$$
is indeed convex. 

While we can solve this without too much trouble (though we need to adjust
our basic gradient descent method to avoid going 'out of bounds'), it turns
out to be easier to use an optimization trick known as *Lagrange Multipliers* 
to change this to something easier to solve. 

The [Method of Lagrange Multipliers](https://en.wikipedia.org/wiki/Lagrange_multiplier)
lets us turn *constrained* optimization problems that look like: 

$$\argmin_x f(x) \text{ such that } g(x) \leq \tau$$

into *unconstrained* problems like: 

$$\argmin_x f(x) + \lambda g(x)$$

Here, the constant $\lambda$ is known as the Lagrange multiplier. Instead of 
_forcing_ $g(x)$ to be bounded, we simply _penalize_ any $x$ that makes $g(\cdot)$
large. For large enough penalties (large $\lambda$), we will eventually force
$g(x)$ to be small enough that we satisfy the original constraint. In fact, 
there is a one-to-one relationship between $\tau$ and $\lambda$, but it's usually
not possible to work it out in any meaningful useful manner; all we can say is that
larger $\lambda$ correspond to smaller $\tau$. 

::: {.callout-note title="Pause to Reflect" collapse="true"}

Why is this true? Why does small $\tau$ imply large $\lambda$ and *vice versa*?

:::

In this form, it's straightforward to pose ridge regression as: 

$$\argmin_{\beta} \frac{1}{2}\|\by - \bX\bbeta\|_2^2 + \frac{\lambda}{2}\|\bbeta\|_2^2$$

::: {.callout-note title="Pause to Reflect" collapse="true"}

Why were we allowed to put an extra $\frac{1}{2}$ in front of the penalty term?

:::

Since both terms of this expression are differentiable, it is easy to show 
that the solution to this problem is given by: 

$$\hat{\bbeta}_{\lambda} = (\bX^{\top}\bX + \lambda \bI)^{-1}\bX^{\top}\by$$

To wit: 

$$\begin{align*}
\bzero &= \frac{\partial}{\partial \bbeta}\left(\frac{1}{2}\|\by - \bX\bbeta\|_2^2 + \frac{\lambda}{2}\|\bbeta\|_2^2\right) \\
       &= -\bX^{\top}(\by - \bX\bbeta) + \lambda \bbeta \\
\bX^{\top}\by &= (\bX^{\top}\bX + \lambda \bI)\bbeta \\
\implies \hat{\beta}_{\text{Ridge}} &= (\bX^{\top}\bX + \lambda \bI)^{-1}\bX^{\top}\by
\end{align*}$$

There are several important things to note about this expression: 

- There are actually *many* ridge regression solutions, one for each value of
  $\lambda$. It is a bit improper to refer to "the" ridge regression solution. 
  We will discuss selecting $\lambda$ below. 
- If you are a bit sloppy, this looks something like: 

  $$\hat{\beta} \approx \frac{SS_{XY}}{SS_{XX} + \lambda}$$
  
  so we 'shrink' the standard OLS solution towards zero, by an amount depending 
  on $\lambda$. Higher $\lambda$ 
- The penalized form (as opposed to the constraint form) allows for $\hat{\beta}$
  to be arbitrarily large, **if the data supports it.**
  
  More than that, $\lambda$ gives us an explicit *trade-off*. When comparing
  two potential solutions ($\hat{\bbeta}_1, \hat{\bbeta}_2$), we will prefer
  $\hat{\bbeta}_1$ if: 
  
  $$\begin{align*}
  \frac{1}{2}\|\by - \bX\bbeta_1\|_2^2 + \frac{\lambda}{2}\|\bbeta_1\|_2^2 & \leq \frac{1}{2}\|\by - \bX\bbeta_2\|_2^2 + \frac{\lambda}{2}\|\bbeta_2\|_2^2 \\
  \|\by - \bX\bbeta_1\|_2^2 - \|\by - \bX\bbeta_2\|_2^2 & \leq \lambda(\|\bbeta_2\|_2^2 - \|\bbeta_1\|_2^2) \\
  \Delta \MSE &\leq \lambda \Delta \text{$\ell_2^2$-Norm}
  \end{align*}$$
  
  That is, we will take a solution with a MSE that is $\lambda$ lower if that lowers
  the squared-norm of the solution by 1. Here $\lambda$ measures the "price" we
  are willing to pay (in terms of MSE) for shrinking the norm of $\hat{\bbeta}$:
  at larger $\lambda$, we demand a *larger* MSE reduction *ceteribus paribus*
  to accept a larger $\hat{\bbeta}$, while at small $\lambda$, we are willing
  to take a larger $\hat{\bbeta}$ in exchange for a much smaller decrease in MSE.
  
- If we set $\lambda = 0$, we recover standard OLS. Continuing the 'price'
  discussion above, this is the limit in which we will take a larger $\hat{\bbeta}$
  for *any* decrease in training MSE. This makes sense as we originally built
  OLS as the ERM for MSE, by definition caring about *only* MSE reduction and
  nothing else. 
- Unlike the OLS solution, the ridge regression exists even when $p > n$ because
  the $+\lambda \bI$ term ensures the 'denominator' prefactor is always 
  invertible.
  
While this isn't as simple to analyze as some of the expressions we considered 
above, this looks similar enough to our 'shrunk' mean estimator that it's 
plausible it will improve on OLS. In fact, you can show that there is always 
_some_ value of $\lambda$ that guarantees improvement over OLS in expectation. 
(This is a result known as the "MSE Existence Theorem")

To wit, 

```{r}
#| warning: false
n <- 100
p <- 80
Z <- matrix(rnorm(n * p), nrow=n)
L <- chol(toeplitz(0.6^(1:p))) # Add 'AR(1)' correlation
X <- Z %*% L
beta <- runif(p, 2, 3)

eye <- function(n) diag(1, n, n)
calculate_ridge_mse <- function(lambda, nreps = 1000){
    MSE <- mean(replicate(nreps, {
        y <- X %*% beta + rnorm(n, sd=1)
        beta_hat <- solve(crossprod(X) + lambda * eye(p), crossprod(X, y))
        sum((beta - beta_hat)^2)
    }))
    
    data.frame(lambda=lambda, MSE=MSE)
}

lambda_grid <- 10^(seq(-2, 2, length.out=41))

RIDGE_MSE <- map(lambda_grid, calculate_ridge_mse) |> list_rbind()
OLS_MSE <- calculate_ridge_mse(0)$MSE # Ridge at lambda = 0 => OLS

ggplot(RIDGE_MSE, aes(x=lambda, y=MSE)) + 
    geom_point() + 
    geom_line() + 
    geom_abline(intercept=OLS_MSE, slope=0, lwd=2, lty=2) + 
    xlab(expression(lambda)) + 
    ylab("Estimation Error (MSE)") + 
    ggtitle("Ridge Regularization Improves on OLS") + 
    theme_bw() + 
    scale_x_log10()
```

Clearly, for smallish $\lambda$, we are actually outperforming OLS, sometimes
by a significant amount! For this problem, the minimum MSE (best estimate) actually
seems to occur near $\lambda \approx$ `r RIDGE_MSE |> filter(MSE == min(MSE)) |> pull(lambda) |> round(3)`
